---
title: "Example code for using R arrow package to handle big dataset"
author: "Zhenguo Zhang"
date: "`r Sys.Date()`"
# use github format
output:
  github_document:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)
workDir <- tempdir()
```

This document is based on the example at https://r4ds.hadley.nz/arrow.

 Apache Arrow, a multi-language toolbox designed for efficient analysis and transport of large datasets. We’ll use Apache Arrow via the arrow package, which provides a dplyr backend allowing you to analyze larger-than-memory datasets using familiar dplyr syntax.

```{r}
library(arrow)
library(tidyverse)
```


## Getting data

### Data

- Content: how many times each book was checked out each month from April 2005 to October 2022
- url:  data.seattle.gov/Community/Checkouts-by-Title/tmmm-ytt6
- size: 41,389,465 rows, 9GB
- format: csv

### Download data

```{r}
dir.create(file.path(workDir, "data"), showWarnings = FALSE)
dataUrl <- "https://r4ds.s3.us-west-2.amazonaws.com/seattle-library-checkouts.csv"
dataFile <- file.path(workDir, "data", "seattle-library-checkouts.csv")
curl::multi_download(dataUrl, destfiles = dataFile, resume = TRUE)
```
### Read data

We will use `arrow::open_dataset()` to read the data, which will 

- scan a few thousand rows to figure out the structure of the dataset
- records what it’s found and stops
- only read further rows as you specifically request them. 

This would avoid reading all the data into memory

```{r}
seattle_csv <- open_dataset(
  sources = file.path(workDir, "data/seattle-library-checkouts.csv"),
  col_types = schema(ISBN = string()),
  format = "csv"
)

seattle_csv
```

As shown in the first line in the output, it tells you that seattle_csv is stored
locally on-disk as a single CSV file.

## Explore data

Let's see what is actually in the dataset

```{r}
seattle_csv |> glimpse()
```

Now let's use dplyr verbs to explore the dataset and use `collect()` to force arrow to perform computation
and return some data.

```{r}
seattle_csv |>
  group_by(CheckoutYear) |>
  summarise(Checkouts = sum(Checkouts)) |>
  arrange(CheckoutYear) |>
  collect()
```


## parquet format

The Parquet format is a columnar storage file format that is optimized for use with big data processing frameworks. It is designed to be efficient in terms of both storage space and query performance.

Like CSV, parquet is used for rectangular data, but instead of being a text format that you can read with any file editor, it’s a custom binary format designed specifically for the needs of big data. This means that:

- Parquet files are usually smaller than the equivalent CSV file, using efficient encodings to keep file size down
  supporting file compression

- Parquet files have a rich type system and store data in a way that records the type along with the data.

- Parquet files are “column-oriented”, much like R’s data frame, leading to better performance for data analysis

- Parquet files are “chunked”, which makes it possible to work on different parts of the file at the same time

### Partitioning

When you have a very large dataset, it’s often useful to split it into multiple files based on the values of
one or more columns (the variables you likely filter on), which will help you to access the data faster.

Arrow suggests that you avoid files smaller than 20MB and larger than 2GB and avoid partitions that produce more than 10,000 files.
 
Now let's rewrite the Seattle library data into different files based on `CheckoutYear`, because it is likely that
some analyses will only want to look at a single year or a few years of data.

```{r}
seattle_csv |>
  write_dataset(
    path = file.path(workDir, "data/seattle_parquet"),
    format = "parquet",
    partitioning = "CheckoutYear"
  )
# alternatively, one can use group_by(CheckoutYear) before write_dataset() to achieve the same effect by replacing the parameter partitioning
```

Let's check the files we just produced

```{r}
tibble(
  files = list.files(
    path = file.path(workDir, "data/seattle_parquet"),
    recursive = TRUE,
    full.names = TRUE
  ),
  size_MB = file.info(files)$size / 2^20
)
```

The file names use a “self-describing” convention used by the [Apache Hive](https://hive.apache.org/) project.
Hive-style partitions name folders with a “key=value” convention.

Another important things is that the total size of the files (around 4GB) is much smaller than the original CSV file.

### Read parquet data

Now we have split the data into multiple parquet files, let's read them back, the syntax
is similar to reading a single file, but the input is the folder containing all the parquet files.

```{r}
seattle_pq <- open_dataset(
  sources = file.path(workDir, "data/seattle_parquet"),
  format = "parquet"
)
print(seattle_pq)
```

Now let's found out the number of books checked each month in the last 5 years:

```{r}
seattle_pq |>
  filter(CheckoutYear >= 2018, MaterialType == "BOOK") |>
  group_by(CheckoutYear, CheckoutMonth) |>
  summarise(TotalCheckouts = sum(Checkouts)) |>
  arrange(CheckoutYear, CheckoutMonth) ->
query
# show what will happen when collect() is called
print(query)
```
And now let's collect the results into R

```{r}
query |> collect()
```

Notes:

- Writing dplyr code for arrow data is conceptually similar to dbplyr: the dplyr code is automatically transformed into a query that the Apache Arrow C++ library understands, which is then executed when you call collect().
- Like dbplyr, arrow only understands some R expressions, so you may not be able to write exactly the same code you usually would. One can find all supported functions in `?acero`

### Compare the performance between single csv file and multiple parquet files

First, let's see how long does it take to calculate the nunber of checkouts in each month of 2021:

```{r}
seattle_csv |>
  filter(CheckoutYear == 2021, MaterialType == "BOOK") |>
  group_by(CheckoutMonth) |>
  summarize(TotalCheckouts = sum(Checkouts)) |>
  arrange(desc(CheckoutMonth)) |>
  collect() |>
  system.time()
```

```{r}
seattle_pq |>
  filter(CheckoutYear == 2021, MaterialType == "BOOK") |>
  group_by(CheckoutMonth) |>
  summarize(TotalCheckouts = sum(Checkouts)) |>
  arrange(desc(CheckoutMonth)) |>
  collect() |>
  system.time()
```

As we can see, reading from multiple parquet files is much faster than reading from a single csv file,
which was achieved for the following reasons:

- arrow is smart enough to recognize that it only needs to read 1 of the 18 parquet files based on
  the filering condition `CheckoutYear == 2021`, while for the csv file, it has to read the whole file

- parquet is binary format, more efficient to read into memory.

- parquet stores data column-wise, and it can only read the four needed columns here in the query:
  `CheckoutYear`, `MaterialType`, `CheckoutMonth`, and `Checkouts`, while for the csv file,
  it has to read all nine columns.


### Challenge examples

Now let's find out the most popular book of each year

```{r}
seattle_pq |>
  filter(MaterialType == "BOOK") |>
  group_by(CheckoutYear, Title) |>
  summarize(TotalCheckouts = sum(Checkouts)) |>
  arrange(CheckoutYear, desc(TotalCheckouts)) |>
  to_duckdb() |> # needs to_duckdb() here to use slice_max()
  slice_max(TotalCheckouts) |>
  collect() |>
  arrange(CheckoutYear)
```

  
## Using duckdb with arrow

[DuckDB](https://duckdb.org/) is an in-process SQL OLAP database management system. It can query various
data sources including parquet files using SQL syntax.  It’s very easy to turn an arrow dataset into a
DuckDB database by calling arrow::to_duckdb(), and we can convert the Seattle parquet dataset into a DuckDB database,
followed by operations, like:

```{r}
seattle_pq |>
  to_duckdb() |>
  filter(CheckoutYear == 2021, MaterialType == "BOOK") |>
  group_by(CheckoutMonth) |>
  summarize(TotalCheckouts = sum(Checkouts)) |>
  arrange(desc(CheckoutMonth)) |>
  collect() |>
  system.time()
```
The nice thing about `to_duckdb()` is that the conversion doesn't involve any memory copying,
enabling seamless transitions from one system to another.

## Different ways of reading data

The `open_dataset()` function is a powerful way to read data, but it’s not the only way.
If you have a single file, you can also use `read_delim_csv()`, `read_csv_arrow()` or `read_parquet()`, 
which will read the entire file into memory as an Arrow Table. This can be useful if you want to
do some in-memory manipulation of the data before using dplyr.

Interestingly, the returned object from `read_csv_arrow()` and `open_dataset()` can be both used with dplyr,
but the former is an Arrow Table, while the latter is an Arrow Dataset. The main difference is that
an Arrow Table is an in-memory data structure that holds the entire dataset, while an Arrow Dataset is a logical representation of a collection of files on disk.

## Interaction with the package `future`

The `future` package provides a simple and consistent API for parallel and distributed computing in R. It allows you to execute code in parallel across multiple cores or machines with different backends, such as multicore, multisession, or cluster.

### The Problem: Non-Exportable Objects

Arrow objects (such as `Table`, `RecordBatch`, `Dataset`, and `RecordBatchReader`) contain external pointers that are tied to the current R session. These objects cannot be directly serialized and sent to worker processes in parallel computing frameworks like `future`. If you try to use Arrow objects directly with `future.apply` or similar functions, you'll encounter errors because the external pointers become invalid in the worker processes.

### Detecting Non-Exportable Objects

You can configure the `future` package to detect non-exportable objects before attempting to export them:

```{r, eval=FALSE}
library(future)
library(future.apply)

# Enable detection of non-exportable objects
options(future.globals.onReference = "error")

# This will fail with an error about external pointers
plan(multisession, workers = 2)
seattle_pq <- open_dataset(
  sources = file.path(workDir, "data/seattle_parquet"),
  format = "parquet"
)

# This would fail:
# future_lapply(1:3, function(x) {
#   seattle_pq |> filter(CheckoutYear == 2020 + x) |> collect()
# })
```

### Solution 1: Marshal and Unmarshal Arrow Objects

The recommended approach is to serialize Arrow objects using `write_to_raw()` and deserialize them in worker processes using `read_ipc_stream()`:

```{r, eval=FALSE}
library(future.apply)
plan(multisession, workers = 2)

# Read data into an Arrow Table
seattle_table <- seattle_pq |>
  filter(CheckoutYear >= 2020) |>
  collect()

# Serialize to raw vector for safe transport
seattle_raw <- write_to_raw(seattle_table, format = "stream")

# Now use it in parallel operations
results <- future_lapply(2020:2022, function(year) {
  # Deserialize in the worker process
  data <- read_ipc_stream(seattle_raw)

  # Perform operations
  data |>
    filter(CheckoutYear == year, MaterialType == "BOOK") |>
    summarise(TotalCheckouts = sum(Checkouts)) |>
    mutate(Year = year)
})

# Combine results
do.call(rbind, results)
```

### Solution 2: Recreate Arrow Objects in Workers

Instead of passing Arrow objects, pass file paths and recreate the objects in each worker:

```{r, eval=FALSE}
plan(multisession, workers = 2)

# Pass the path, not the object
data_path <- file.path(workDir, "data/seattle_parquet")

results <- future_lapply(2020:2022, function(year) {
  # Recreate the dataset in each worker
  pq <- arrow::open_dataset(data_path, format = "parquet")

  pq |>
    filter(CheckoutYear == year, MaterialType == "BOOK") |>
    group_by(CheckoutMonth) |>
    summarise(TotalCheckouts = sum(Checkouts)) |>
    collect() |>
    mutate(Year = year)
}, future.packages = "arrow")

# Combine results
bind_rows(results)
```

### Solution 3: Use Arrow's Built-in Parallelism

For many operations, Arrow already provides parallel execution through its C++ compute engine. You may not need `future` at all:

```{r, eval=FALSE}
# Control the number of CPU threads Arrow uses
set_cpu_count(4)

# Arrow automatically uses multiple threads for operations
seattle_pq |>
  filter(MaterialType == "BOOK") |>
  group_by(CheckoutYear) |>
  summarise(TotalCheckouts = sum(Checkouts)) |>
  collect()
```

### Best Practices

1. **Prefer Arrow's native parallelism** for operations on Arrow datasets, as it's highly optimized
2. **Use file paths** instead of objects when possible with `future`
3. **Marshal objects with `write_to_raw()`** when you need to pass data to workers
4. **Specify `future.packages = "arrow"`** to ensure the arrow package is loaded in workers
5. **Use `future.seed = TRUE`** for reproducible random operations across workers

### Example: Parallel Processing of Multiple Datasets

```{r, eval=FALSE}
plan(multisession, workers = 2)

# Process multiple years in parallel
years_to_process <- 2018:2022
data_path <- file.path(workDir, "data/seattle_parquet")

results <- future_lapply(years_to_process, function(year) {
  library(arrow)
  library(dplyr)

  dataset <- open_dataset(data_path, format = "parquet")

  dataset |>
    filter(CheckoutYear == year) |>
    group_by(MaterialType) |>
    summarise(
      TotalCheckouts = sum(Checkouts),
      UniqueItems = n_distinct(Title)
    ) |>
    collect() |>
    mutate(Year = year)
}, future.packages = c("arrow", "dplyr"), future.seed = TRUE)

# Combine and analyze
final_result <- bind_rows(results)
print(final_result)
```



## Resources

- parquet format: https://parquet.apache.org/

- arrow package: https://arrow.apache.org/

