---
title: "ggiraph: Interactive ggplot2 Graphics"
author: "Tutorial"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

## What is ggiraph?

**ggiraph** is an R package that transforms static ggplot2 graphics into interactive visualizations for web browsers. It extends ggplot2 by adding interactivity capabilities such as tooltips, hover effects, click actions, and element selection.

## Main Use Cases

- **Data exploration**: Add tooltips to display detailed information on hover
- **Presentations**: Create engaging, interactive visualizations for reports and dashboards
- **Shiny applications**: Enable user interaction with plot elements for filtering and selection
- **Web-based reports**: Publish interactive graphics in R Markdown documents

## Why Use ggiraph?

- Seamless integration with ggplot2 - minimal code changes required
- No JavaScript knowledge needed
- Works in R Markdown, Shiny, and standalone HTML
- Supports almost all ggplot2 geometries and themes
- Lightweight and fast rendering

## Prerequisites

- Basic knowledge of R and ggplot2
- R version 3.5.0 or higher recommended
- Familiarity with ggplot2 syntax and aesthetics

# Installation & Setup

## Install from CRAN

```{r install-cran, eval=FALSE}
install.packages("ggiraph")
```

## Install Development Version

For the latest features and bug fixes:

```{r install-dev, eval=FALSE}
# Install devtools if not already installed
install.packages("devtools")

# Install ggiraph from GitHub
devtools::install_github('davidgohel/ggiraph')
```

## Load the Package

```{r load-packages}
library(ggplot2)
library(ggiraph)
library(dplyr)  # for data manipulation
```

## Verify Installation

Create a simple interactive plot to verify the installation:

```{r verify-install}
# Simple example
data <- data.frame(
  x = 1:5,
  y = 1:5,
  label = paste("Point", 1:5)
)

gg <- ggplot(data, aes(x = x, y = y)) +
  geom_point_interactive(aes(tooltip = label), size = 4) +
  theme_minimal()

girafe(ggobj = gg)
```

If you see an interactive plot with tooltips appearing on hover, the installation was successful!

# Key Concepts

## The Three Special Interactive Aesthetics

ggiraph introduces three special aesthetics that add interactivity to ggplot2 graphics:

| Aesthetic | Description | Example Use Case |
|-----------|-------------|------------------|
| `tooltip` | Text displayed when hovering over an element | Show detailed data values, labels, or descriptions |
| `data_id` | Identifier that links elements together | Highlight related elements across multiple plots or layers |
| `onclick` | JavaScript code executed when element is clicked | Open URLs, trigger custom actions, navigate to external resources |

### How Interactive Aesthetics Work

These aesthetics are added inside the `aes()` function of interactive geometries:

```{r interactive-aesthetics-demo, eval=FALSE}
geom_point_interactive(aes(
  x = x_var,
  y = y_var,
  tooltip = "Information to display",      # Tooltip text
  data_id = "unique_identifier",           # Grouping ID
  onclick = "window.open('http://url')"    # JavaScript action
))
```

## Interactive Geometries

ggiraph provides interactive versions of ggplot2 geometries by adding the `_interactive` suffix:

| Standard ggplot2 | ggiraph Interactive | Purpose |
|------------------|---------------------|---------|
| `geom_point()` | `geom_point_interactive()` | Interactive scatter plots |
| `geom_bar()` | `geom_bar_interactive()` | Interactive bar charts |
| `geom_col()` | `geom_col_interactive()` | Interactive column charts |
| `geom_line()` | `geom_line_interactive()` | Interactive line plots |
| `geom_text()` | `geom_text_interactive()` | Interactive text labels |
| `geom_polygon()` | `geom_polygon_interactive()` | Interactive polygons/maps |
| `geom_tile()` | `geom_tile_interactive()` | Interactive heatmaps |

## The girafe() Function

The `girafe()` function converts a ggplot object into an interactive graphic:

```{r girafe-concept, eval=FALSE}
# Basic usage
girafe(ggobj = your_ggplot_object)

# With customization
girafe(
  ggobj = your_ggplot_object,
  width_svg = 7,      # Width in inches
  height_svg = 5,     # Height in inches
  options = list(...)  # Interactive options
)
```

# Key Functions and Commands

## Core Functions

| Function | Purpose | Example |
|----------|---------|---------|
| `girafe()` | Create interactive graphic from ggplot object | `girafe(ggobj = p)` |
| `girafe_options()` | Customize interactive behaviors | `girafe_options(x, opts_hover(...))` |
| `geom_*_interactive()` | Interactive geometry functions | `geom_point_interactive(aes(tooltip = label))` |

## Customization Functions (opts_* family)

| Function | Purpose | Example |
|----------|---------|---------|
| `opts_tooltip()` | Customize tooltip appearance | `opts_tooltip(opacity = 0.8, css = "...")` |
| `opts_hover()` | Define hover effects | `opts_hover(css = "fill:red;stroke:black;")` |
| `opts_selection()` | Control selection behavior | `opts_selection(type = "single", css = "...")` |
| `opts_zoom()` | Configure zoom capabilities | `opts_zoom(min = 0.5, max = 4)` |
| `opts_toolbar()` | Show/hide interactive toolbar | `opts_toolbar(position = "topright")` |

## Shiny Integration Functions

| Function | Purpose | Example |
|----------|---------|---------|
| `girafeOutput()` | Create girafe output element in Shiny UI | `girafeOutput("plot1")` |
| `renderGirafe()` | Render reactive girafe in Shiny server | `renderGirafe({ girafe(...) })` |

# Full Working Examples

## Example 1: Basic Interactive Scatter Plot with Tooltips

```{r example1}
# Prepare data
data <- mtcars
data$carname <- rownames(mtcars)

# Create ggplot with interactive points
gg_scatter <- ggplot(data, aes(x = wt, y = mpg)) +
  geom_point_interactive(
    aes(tooltip = carname, data_id = carname),
    size = 3,
    color = "steelblue"
  ) +
  labs(
    title = "Car Weight vs. Fuel Efficiency",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_minimal()

# Convert to interactive graphic
girafe(ggobj = gg_scatter)
```

## Example 2: Color-Coded Plot with Rich Tooltips

```{r example2}
# Create rich tooltip text
data$tooltip_text <- paste0(
  "<b>", data$carname, "</b><br/>",
  "MPG: ", data$mpg, "<br/>",
  "Weight: ", data$wt, " (1000 lbs)<br/>",
  "Cylinders: ", data$cyl
)

# Create plot with color coding
gg_color <- ggplot(data, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point_interactive(
    aes(tooltip = tooltip_text, data_id = carname),
    size = 4
  ) +
  scale_color_brewer(palette = "Set1", name = "Cylinders") +
  labs(
    title = "Car Performance by Cylinder Count",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_minimal()

# Add interactive options
girafe(ggobj = gg_color)
```

## Example 3: Interactive Bar Chart with Click Actions

```{r example3}
# Aggregate data
cyl_summary <- data %>%
  group_by(cyl) %>%
  summarise(
    avg_mpg = mean(mpg),
    count = n()
  ) %>%
  mutate(
    tooltip = paste0("Cylinders: ", cyl, "\nAvg MPG: ", round(avg_mpg, 1), "\nCount: ", count),
    # Create search URL for each cylinder count
    onclick = paste0("window.open('https://www.google.com/search?q=", cyl, "+cylinder+cars')")
  )

# Create interactive bar chart
gg_bar <- ggplot(cyl_summary, aes(x = factor(cyl), y = avg_mpg)) +
  geom_col_interactive(
    aes(tooltip = tooltip, onclick = onclick, data_id = cyl),
    fill = "steelblue"
  ) +
  labs(
    title = "Average MPG by Cylinder Count (Click to Search)",
    x = "Number of Cylinders",
    y = "Average MPG"
  ) +
  theme_minimal()

girafe(ggobj = gg_bar)
```

## Example 4: Advanced Customization with Hover and Selection Effects

```{r example4}
# Create plot with interactive features
gg_advanced <- ggplot(data, aes(x = hp, y = mpg, size = wt)) +
  geom_point_interactive(
    aes(
      tooltip = carname,
      data_id = carname,
      color = factor(cyl)
    ),
    alpha = 0.7
  ) +
  scale_color_brewer(palette = "Dark2", name = "Cylinders") +
  scale_size_continuous(name = "Weight", range = c(2, 10)) +
  labs(
    title = "Horsepower vs. Efficiency (Hover and Click)",
    x = "Horsepower",
    y = "Miles per Gallon"
  ) +
  theme_minimal()

# Add extensive customization
girafe(
  ggobj = gg_advanced,
  width_svg = 8,
  height_svg = 6,
  options = list(
    opts_hover(css = "fill:orange;stroke:black;stroke-width:2px;"),
    opts_selection(
      type = "multiple",
      css = "fill:red;stroke:black;stroke-width:3px;"
    ),
    opts_tooltip(
      opacity = 0.9,
      css = "background-color:white;padding:10px;border-radius:5px;box-shadow:2px 2px 5px gray;"
    ),
    opts_zoom(min = 0.5, max = 3),
    opts_toolbar(position = "topright")
  )
)
```

## Example 5: Multiple Linked Plots with Shared data_id

```{r example5, fig.width=10}
# Create two plots that share data_id for linked highlighting
data$tooltip_simple <- paste0(data$carname, "\nMPG: ", data$mpg)

# Plot 1: Weight vs MPG
p1 <- ggplot(data, aes(x = wt, y = mpg)) +
  geom_point_interactive(
    aes(tooltip = tooltip_simple, data_id = carname),
    size = 3, color = "steelblue"
  ) +
  labs(title = "Weight vs MPG", x = "Weight", y = "MPG") +
  theme_minimal()

# Plot 2: Horsepower vs MPG
p2 <- ggplot(data, aes(x = hp, y = mpg)) +
  geom_point_interactive(
    aes(tooltip = tooltip_simple, data_id = carname),
    size = 3, color = "darkgreen"
  ) +
  labs(title = "Horsepower vs MPG", x = "HP", y = "MPG") +
  theme_minimal()

# Combine plots using patchwork or cowplot (if available)
# For this example, we'll show them separately
girafe(
  ggobj = p1,
  options = list(
    opts_hover(css = "fill:orange;stroke:black;r:6pt;")
  )
)

girafe(
  ggobj = p2,
  options = list(
    opts_hover(css = "fill:orange;stroke:black;r:6pt;")
  )
)
```

**Note**: When using the same `data_id` across multiple girafe plots in a single HTML document, hovering over an element in one plot can highlight the corresponding element in other plots (when properly configured in the document).

# Common Patterns & Best Practices

## 1. Creating Informative Tooltips

**Best Practice**: Use HTML formatting for rich, multi-line tooltips:

```{r tooltip-pattern, eval=FALSE}
# Create formatted tooltip with HTML
data$tooltip <- paste0(
  "<b style='color:#333;'>", data$name, "</b><br/>",
  "<i>Category:</i> ", data$category, "<br/>",
  "<i>Value:</i> ", round(data$value, 2)
)

geom_point_interactive(aes(tooltip = tooltip))
```

## 2. Handling Large Datasets

For datasets with many points, consider:

- Using `alpha` transparency to reduce visual clutter
- Limiting tooltip text length
- Using `opts_hover(css = "...")` to make hover effects subtle

```{r large-data-pattern, eval=FALSE}
ggplot(large_data, aes(x = x, y = y)) +
  geom_point_interactive(
    aes(tooltip = short_label),
    size = 2,
    alpha = 0.5  # Add transparency
  )
```

## 3. Consistent Styling Across Plots

Define reusable option lists:

```{r styling-pattern, eval=FALSE}
# Define standard options
my_options <- list(
  opts_hover(css = "fill:orange;stroke:black;stroke-width:2px;"),
  opts_tooltip(
    opacity = 0.9,
    css = "background-color:white;padding:8px;border-radius:4px;"
  )
)

# Reuse across multiple plots
girafe(ggobj = plot1, options = my_options)
girafe(ggobj = plot2, options = my_options)
```

## 4. Grouping with data_id

**Important**: Interactive aesthetics can affect ggplot2's automatic grouping. When using `data_id`, explicitly specify the `group` aesthetic if needed:

```{r grouping-pattern, eval=FALSE}
ggplot(data, aes(x = time, y = value, color = category)) +
  geom_line_interactive(
    aes(
      tooltip = label,
      data_id = category,
      group = category  # Explicit grouping
    )
  )
```

## 5. Testing Interactivity

Always test your interactive plots in the target environment:

- **R Markdown**: Knit the document to HTML
- **Shiny**: Run the app in a browser (not RStudio Viewer for full functionality)
- **Standalone HTML**: Save using `htmlwidgets::saveWidget()`

## Common Pitfalls to Avoid

1. **Forgetting to use `girafe()`**: Interactive geoms alone don't create interactivity - you must wrap with `girafe()`
2. **Overly complex tooltips**: Keep tooltips concise for better user experience
3. **Inconsistent `data_id`**: Ensure matching elements have identical `data_id` values
4. **Not testing in browser**: Some features work differently in RStudio Viewer vs. browser

# References & Further Reading

## Official Documentation

- **Main Website**: https://davidgohel.github.io/ggiraph/
- **Function Reference**: https://davidgohel.github.io/ggiraph/reference/index.html
- **GitHub Repository**: https://github.com/davidgohel/ggiraph

## Key Documentation Pages Used

1. ggiraph Homepage - Overview and installation
2. Function Reference - Complete list of functions and their parameters
3. girafe() Documentation - Core function details
4. girafe_options() Documentation - Customization options

## Getting Help

- **Stack Overflow**: Use tags `ggiraph` and `r` for questions
- **GitHub Issues**: Report bugs at https://github.com/davidgohel/ggiraph/issues
- **RStudio Community**: Search for ggiraph discussions

## Related Packages

- **plotly**: Alternative for interactive ggplot2 graphics with different features
- **htmlwidgets**: Underlying framework for R/HTML interactivity
- **shiny**: Build interactive web applications with ggiraph
- **patchwork**: Combine multiple ggiraph plots in layouts

## Additional Learning Resources

- ggplot2 documentation: https://ggplot2.tidyverse.org/
- HTML/CSS basics for customizing tooltips and hover effects
- JavaScript basics for advanced `onclick` actions

---

**Tutorial created**: `r Sys.Date()`
**ggiraph version used**: `r packageVersion("ggiraph")`
