---
title: "Learning Patterns in ggplot2: ggpattern and fillpattern"
author: "Learning Notes"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
```

# Overview

This document explores two packages for adding patterns to ggplot2 visualizations:

- **ggpattern**: Feature-rich package with extensive pattern types and customization
- **fillpattern**: Simpler, more performant alternative with basic pattern support

**Performance Note**: ggpattern can be quite slow, especially with complex patterns or large datasets, due to its grid-based rendering implementation. For production visualizations with large data, consider fillpattern or pre-generating pattern images.

Focus areas:
1. Using patterns as aesthetics
2. Controlling pattern shapes, densities, and appearance
3. Ensuring patterns appear properly in legend keys

# Setup

```{r install-packages, eval=FALSE}
# Install packages if needed
install.packages("ggpattern")
install.packages("fillpattern")
```

```{r load-libraries}
library(ggplot2)
library(ggpattern)
library(dplyr)
```

---

# Part 1: Pattern Categories in ggpattern

ggpattern organizes patterns into three main categories based on their rendering approach:

## 1.1 Stripe Patterns

**Pattern types**: `stripe`, `crosshatch`, `wave`, `weave`

These are line-based patterns with regular spacing. They are the most commonly used and relatively performant.

**Key parameters**:
- `pattern_density`: Controls how many lines appear (higher = more lines)
- `pattern_spacing`: Distance between pattern elements (lower = denser)
- `pattern_angle`: Rotation angle in degrees (0 = vertical, 90 = horizontal, 45 = diagonal)

```{r stripe-patterns}
df_stripes <- data.frame(
  pattern_type = c("stripe", "crosshatch", "wave", "weave"),
  value = c(10, 10, 10, 10)
)

ggplot(df_stripes, aes(x = pattern_type, y = value)) +
  geom_col_pattern(
    aes(pattern = pattern_type),
    fill = "lightblue",
    colour = "black",
    pattern_fill = "navy",
    pattern_spacing = 0.025,
    pattern_density = 0.5,
    pattern_key_scale_factor = 1.5
  ) +
  scale_pattern_manual(values = c(crossbatch = "crosshatch", stripe = "stripe", wave = "wave", weave = "weave")) +
  theme_minimal() +
  labs(
    title = "Stripe-Based Patterns",
    subtitle = "Regular line-based patterns"
  )
```

### Controlling Stripe Orientation and Density

```{r stripe-control}
df_angles <- data.frame(
  group = c(
    "Vertical\n(0°)", "Horizontal\n(90°)", "Diagonal\n(45°)",
    "Dense\n(spacing=0.01)", "Sparse\n(spacing=0.05)"
  ),
  angle = c(0, 90, 45, 45, 45),
  spacing = c(0.025, 0.025, 0.025, 0.01, 0.05),
  value = c(10, 10, 10, 10, 10)
)

ggplot(df_angles, aes(x = group, y = value)) +
  geom_col_pattern(
    aes(pattern_angle = angle, pattern_spacing = spacing),
    pattern = "stripe",
    pattern_fill = "darkred",
    fill = "pink",
    colour = "black",
    pattern_key_scale_factor = 1.5
  ) +
  theme_minimal() +
  labs(
    title = "Controlling Stripe Angle and Spacing",
    subtitle = "Lower spacing = denser patterns"
  ) +
  theme(axis.text.x = element_text(size = 9))
```

## 1.2 Geometric Patterns

**Pattern types**: `circle`, `regular_polygon`, `rose`, `pch`

These patterns fill the space with repeated geometric shapes.

**Key parameters**:
- `pattern_size`: Size of individual pattern elements (larger = bigger shapes)
- `pattern_spacing`: Distance between shapes (lower = more tightly packed)
- `pattern_shape`: For `regular_polygon`, number of sides (3=triangle, 4=square, etc.)
- `pattern_density`: Controls how many shapes appear per unit area

```{r geometric-patterns}
df_geom <- data.frame(
  pattern_type = c("circle", "regular_polygon", "rose"),
  value = c(10, 10, 10)
)

ggplot(df_geom, aes(x = pattern_type, y = value)) +
  geom_col_pattern(
    aes(pattern = pattern_type),
    fill = "lightyellow",
    colour = "black",
    pattern_fill = "darkgreen",
    pattern_size = 10,
    pattern_spacing = 0.03,
    pattern_shape = 6, # hexagon for regular_polygon (ignored for other patterns)
    pattern_key_scale_factor = 2.0
  ) +
  scale_pattern_manual(
    values = c("circle", "regular_polygon", "rose")
  ) +
  theme_minimal() +
  labs(
    title = "Geometric Patterns",
    subtitle = "Shape-based patterns (hexagon shown for regular_polygon)"
  )
```

### Controlling Shape Size and Density

```{r geometric-size}
df_size <- data.frame(
  label = c(
    "Small\n(size=5)", "Medium\n(size=10)", "Large\n(size=15)",
    "Dense\n(spacing=0.02)", "Sparse\n(spacing=0.05)"
  ),
  size = c(5, 10, 15, 10, 10),
  spacing = c(0.03, 0.03, 0.03, 0.02, 0.05),
  value = c(10, 10, 10, 10, 10)
)

ggplot(df_size, aes(x = label, y = value)) +
  geom_col_pattern(
    aes(pattern_size = size, pattern_spacing = spacing),
    pattern = "circle",
    pattern_fill = "purple",
    fill = "lavender",
    colour = "black",
    pattern_key_scale_factor = 1.8
  ) +
  theme_minimal() +
  labs(
    title = "Controlling Geometric Pattern Size and Spacing",
    subtitle = "Larger size = bigger shapes; smaller spacing = denser packing"
  ) +
  theme(axis.text.x = element_text(size = 9))
```

## 1.3 Noise/Gradient Patterns

**Pattern types**: `gradient`, `plasma`, `ambient`

These are algorithmic patterns that create smooth, noise-based fills. They can be quite slow to render.

**Key parameters**:
- `pattern_fill`, `pattern_fill2`: Colors for gradient endpoints
- `pattern_orientation`: Direction of gradient
- `pattern_frequency`: For plasma/ambient, controls noise frequency

```{r noise-patterns}
df_noise <- data.frame(
  pattern_type = c("gradient", "plasma"),
  value = c(10, 10)
)

ggplot(df_noise, aes(x = pattern_type, y = value)) +
  geom_col_pattern(
    aes(pattern = pattern_type),
    fill = "white",
    colour = "black",
    pattern_fill = "blue",
    pattern_fill2 = "red",
    pattern_key_scale_factor = 1.5
  ) +
  scale_pattern_manual(values = c("gradient", "plasma")) +
  theme_minimal() +
  labs(
    title = "Noise/Gradient Patterns",
    subtitle = "Algorithmic patterns (slow to render)"
  )
```

---

# Part 2: Using Patterns as Aesthetics

## 2.1 Mapping Patterns to Categorical Variables

Just like you map `fill` or `color` to variables, you can map `pattern` to distinguish groups.

```{r pattern-mapping}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot_pattern(
    aes(pattern = factor(cyl)),
    pattern_fill = "gray30",
    fill = "white",
    pattern_spacing = 0.025,
    pattern_key_scale_factor = 1.5
  ) +
  scale_pattern_manual(
    name = "Cylinders",
    values = c("stripe", "crosshatch", "circle")
  ) +
  theme_minimal() +
  labs(
    title = "Patterns Mapped to Cylinder Count",
    subtitle = "Different pattern types for each group"
  )
```

## 2.2 Combining Pattern and Fill Aesthetics

You can map both `fill` color and `pattern` type to different variables for multi-dimensional encoding.

```{r pattern-fill-combo}
ggplot(mtcars, aes(x = factor(cyl), fill = factor(am))) +
  geom_bar_pattern(
    aes(pattern = factor(am)),
    position = "dodge",
    colour = "black",
    pattern_fill = "gray20",
    pattern_spacing = 0.025,
    pattern_key_scale_factor = 1.5
  ) +
  scale_pattern_manual(
    name = "Transmission",
    values = c("stripe", "crosshatch"),
    labels = c("Automatic", "Manual")
  ) +
  scale_fill_manual(
    name = "Transmission",
    values = c("steelblue", "coral"),
    labels = c("Automatic", "Manual")
  ) +
  theme_minimal() +
  labs(
    title = "Pattern Type and Fill Color by Transmission",
    subtitle = "Two aesthetics encoding the same variable"
  )
```

## 2.3 Mapping Multiple Pattern Parameters

You can map multiple pattern aesthetics (density, angle, type) to variables for fine control.

```{r multiple-pattern-aes}
ggplot(mtcars, aes(x = factor(cyl), fill = factor(gear))) +
  geom_bar_pattern(
    aes(
      pattern = factor(gear),
      pattern_angle = as.numeric(factor(gear)) * 30, # 30°, 60°, 90°
      pattern_density = as.numeric(factor(gear)) * 0.15 # 0.15, 0.3, 0.45
    ),
    position = "dodge",
    colour = "black",
    pattern_fill = "gray30",
    pattern_key_scale_factor = 1.8
  ) +
  scale_pattern_manual(
    name = "Gears",
    values = c("stripe", "crosshatch", "circle")
  ) +
  scale_fill_manual(
    name = "Gears",
    values = c("#E69F00", "#56B4E9", "#009E73")
  ) +
  theme_minimal() +
  labs(
    title = "Multiple Pattern Aesthetics",
    subtitle = "Varying pattern type, angle, and density by gear count"
  )
```

---

# Part 3: Key Pattern Parameters

## 3.1 Core Pattern Control Parameters

| Parameter | Effect When Increased | Typical Range | Notes |
|-----------|----------------------|---------------|-------|
| `pattern_density` | More pattern elements appear, making pattern denser | 0.1 - 1.0 | Higher values = more crowded patterns |
| `pattern_spacing` | Pattern elements spread farther apart, making pattern sparser | 0.01 - 0.1 | **Lower values = denser** (inverse relationship) |
| `pattern_angle` | Rotates pattern clockwise by specified degrees | 0 - 360 | 0=vertical, 45=diagonal, 90=horizontal |
| `pattern_size` | Individual pattern elements become larger | 1 - 20 | Applies to geometric patterns (circles, polygons) |
| `pattern_fill` | Changes the color of the pattern elements themselves | Any color | The "ink" color of the pattern |
| `pattern_colour` | Changes the outline/stroke color of pattern elements | Any color | Use British spelling `colour` |

### Important Distinction: density vs spacing

- **`pattern_density`**: Controls how much of the fill area is covered by the pattern (higher = more coverage)
- **`pattern_spacing`**: Controls distance between elements (lower = elements closer together)

For stripe patterns, reducing `pattern_spacing` makes stripes closer together (denser). Increasing `pattern_density` adds more stripes overall.

```{r density-vs-spacing}
df_compare <- data.frame(
  type = rep(c("Density\nControl", "Spacing\nControl"), each = 3),
  level = rep(c("Low", "Medium", "High"), 2),
  density = c(0.2, 0.5, 0.8, 0.5, 0.5, 0.5),
  spacing = c(0.03, 0.03, 0.03, 0.05, 0.03, 0.015),
  x = 1:6
)

ggplot(df_compare, aes(x = factor(x), y = 10)) +
  geom_col_pattern(
    aes(pattern_density = density, pattern_spacing = spacing),
    pattern = "stripe",
    pattern_fill = "navy",
    fill = "lightblue",
    colour = "black",
    pattern_key_scale_factor = 1.5
  ) +
  scale_x_discrete(labels = paste(df_compare$type, df_compare$level, sep = "\n")) +
  theme_minimal() +
  labs(
    title = "Pattern Density vs Spacing",
    subtitle = "Density: 0.2→0.5→0.8 | Spacing: 0.05→0.03→0.015 (note inverse)",
    x = NULL,
    y = "Value"
  ) +
  theme(axis.text.x = element_text(size = 8))
```

## 3.2 Pattern Shape Control (for regular_polygon)

The `pattern_shape` parameter controls the number of sides for `regular_polygon` patterns.

```{r polygon-shapes}
df_shapes <- data.frame(
  shape_name = c("Triangle", "Square", "Pentagon", "Hexagon", "Octagon"),
  n_sides = c(3, 4, 5, 6, 8),
  value = rep(10, 5)
)

ggplot(df_shapes, aes(x = shape_name, y = value)) +
  geom_col_pattern(
    aes(pattern_shape = factor(n_sides)),
    pattern = "regular_polygon",
    pattern_fill = "darkblue",
    fill = "lightblue",
    colour = "black",
    pattern_size = 12,
    pattern_spacing = 0.03,
    pattern_key_scale_factor = 2.0
  ) +
  scale_pattern_shape_manual(values = c(3, 4, 5, 6, 8)) +
  theme_minimal() +
  labs(
    title = "Regular Polygon Pattern Shapes",
    subtitle = "pattern_shape controls number of sides (3-8 shown)"
  )
```

---

# Part 4: Legend Key Control

## 4.1 The Critical Parameter: pattern_key_scale_factor

**Problem**: By default, patterns in legend keys are often too small or too sparse to be visible.

**Solution**: Use `pattern_key_scale_factor` to scale up the pattern appearance in legends without affecting the plot.

```{r legend-comparison}
# Without pattern_key_scale_factor (hard to see)
p1 <- ggplot(mtcars, aes(x = factor(cyl), fill = factor(gear))) +
  geom_bar_pattern(
    aes(pattern = factor(gear)),
    position = "dodge",
    colour = "black",
    pattern_spacing = 0.025,
    pattern_fill = "gray30"
  ) +
  scale_pattern_manual(
    name = "Gears",
    values = c("stripe", "crosshatch", "circle")
  ) +
  scale_fill_manual(
    name = "Gears",
    values = c("coral", "steelblue", "lightgreen")
  ) +
  theme_minimal() +
  labs(title = "Default Legend (patterns barely visible)")

# With pattern_key_scale_factor (much clearer)
p2 <- ggplot(mtcars, aes(x = factor(cyl), fill = factor(gear))) +
  geom_bar_pattern(
    aes(pattern = factor(gear)),
    position = "dodge",
    colour = "black",
    pattern_spacing = 0.025,
    pattern_fill = "gray30",
    pattern_key_scale_factor = 1.8 # Makes patterns visible in legend
  ) +
  scale_pattern_manual(
    name = "Gears",
    values = c("stripe", "crosshatch", "circle")
  ) +
  scale_fill_manual(
    name = "Gears",
    values = c("coral", "steelblue", "lightgreen")
  ) +
  theme_minimal() +
  labs(title = "With pattern_key_scale_factor = 1.8")

p1
p2
```

### How pattern_key_scale_factor Works

**What it does**: Multiplies the density/frequency of pattern elements specifically in the legend keys.

- **Value = 1.0**: Default (no scaling)
- **Value = 1.5-2.0**: Recommended for most patterns (50-100% more dense in legend)
- **Value > 2.0**: For very sparse patterns or geometric patterns

**Note**: This parameter only affects the legend appearance, not the plot itself.

## 4.2 Combining with theme() for Perfect Legends

For optimal legend visibility, combine `pattern_key_scale_factor` with `theme()` settings.

```{r perfect-legend}
ggplot(mtcars, aes(x = factor(cyl), fill = factor(am))) +
  geom_boxplot_pattern(
    aes(pattern = factor(am), pattern_angle = factor(am)),
    colour = "black",
    pattern_fill = "gray20",
    pattern_spacing = 0.025,
    pattern_key_scale_factor = 2.0 # Double density in legend
  ) +
  scale_pattern_manual(
    name = "Transmission",
    values = c("stripe", "crosshatch"),
    labels = c("Automatic", "Manual")
  ) +
  scale_pattern_angle_manual(
    name = "Transmission",
    values = c(45, 0),
    labels = c("Automatic", "Manual")
  ) +
  scale_fill_manual(
    name = "Transmission",
    values = c("coral", "steelblue"),
    labels = c("Automatic", "Manual")
  ) +
  theme_minimal() +
  theme(
    legend.key.size = unit(1.5, "cm"), # Larger legend boxes
    legend.key = element_rect(colour = "black"), # Border around keys
    legend.position = "right"
  ) +
  labs(
    title = "Perfect Pattern Legends",
    subtitle = "pattern_key_scale_factor + large legend.key.size",
    x = "Cylinders",
    y = "MPG"
  )
```

---

# Part 5: Advanced Techniques

## 5.1 Pattern Fill vs Base Fill

Understanding the relationship between `fill` and `pattern_fill`:

- **`fill`**: Background color of the entire geom area
- **`pattern_fill`**: Color of the pattern elements themselves

```{r fill-vs-pattern-fill}
df_fills <- data.frame(
  combo = c(
    "Light bg\nDark pattern", "Dark bg\nLight pattern",
    "White bg\nColored pattern", "Colored bg\nWhite pattern"
  ),
  fill_color = c("lightblue", "navy", "white", "coral"),
  pattern_color = c("navy", "lightblue", "red", "white"),
  value = c(10, 10, 10, 10)
)

ggplot(df_fills, aes(x = combo, y = value)) +
  geom_col_pattern(
    aes(fill = fill_color, pattern_fill = pattern_color),
    pattern = "stripe",
    pattern_spacing = 0.025,
    colour = "black",
    pattern_key_scale_factor = 1.5
  ) +
  scale_fill_identity() +
  scale_pattern_fill_identity() +
  theme_minimal() +
  labs(
    title = "Base Fill vs Pattern Fill",
    subtitle = "Contrasting colors improve pattern visibility"
  )
```

## 5.2 Using Patterns for Accessibility

Patterns are valuable for creating colorblind-friendly visualizations. Combine with colorblind-safe palettes.

```{r colorblind-friendly}
# Colorblind-safe palette + patterns
ggplot(mtcars, aes(x = factor(cyl), fill = factor(gear))) +
  geom_bar_pattern(
    aes(pattern = factor(gear)),
    position = "dodge",
    colour = "black",
    pattern_fill = "gray30",
    pattern_spacing = 0.025,
    pattern_key_scale_factor = 1.8
  ) +
  scale_pattern_manual(
    name = "Gears",
    values = c("stripe", "crosshatch", "none")
  ) +
  scale_fill_manual(
    name = "Gears",
    values = c("#E69F00", "#56B4E9", "#009E73") # Colorblind-safe
  ) +
  theme_minimal() +
  labs(
    title = "Accessible Visualization with Patterns",
    subtitle = "Patterns + colorblind-safe colors for maximum accessibility"
  )
```

## 5.3 Complex Multi-Parameter Patterns

```{r complex-example}
# Create sample data with multiple grouping variables
df_complex <- mtcars %>%
  mutate(
    cyl_cat = factor(cyl),
    am_cat = factor(am, labels = c("Auto", "Manual")),
    gear_cat = factor(gear)
  ) %>%
  group_by(cyl_cat, am_cat) %>%
  summarise(mean_mpg = mean(mpg), .groups = "drop")

ggplot(df_complex, aes(x = cyl_cat, y = mean_mpg, fill = am_cat)) +
  geom_col_pattern(
    aes(
      pattern = am_cat,
      pattern_angle = as.numeric(am_cat) * 45,
      pattern_spacing = 0.02 + as.numeric(am_cat) * 0.01
    ),
    position = "dodge",
    colour = "black",
    pattern_fill = "gray20",
    pattern_key_scale_factor = 2.0
  ) +
  scale_pattern_manual(
    name = "Transmission",
    values = c("stripe", "crosshatch")
  ) +
  scale_fill_manual(
    name = "Transmission",
    values = c("#1b9e77", "#d95f02")
  ) +
  theme_minimal() +
  theme(
    legend.key.size = unit(1.3, "cm"),
    legend.position = "right"
  ) +
  labs(
    title = "Average MPG by Cylinders and Transmission",
    subtitle = "Multiple pattern parameters for rich encoding",
    x = "Cylinders",
    y = "Mean MPG"
  )
```

---

# Part 6: fillpattern Package

## 6.1 Introduction to fillpattern

`fillpattern` is a lighter-weight alternative to `ggpattern` with:
- **Better performance**: Faster rendering for large datasets
- **Simpler API**: Fewer pattern types and options
- **Different approach**: Uses grid patterns rather than ggpattern's rendering engine

## 6.2 Basic fillpattern Usage

```{r fillpattern-basic}
library(fillpattern)

# Sample data
df <- data.frame(
  category = LETTERS[1:4],
  value = c(10, 25, 15, 30)
)

# Basic pattern with fillpattern
# Note: fillpattern uses different pattern names than ggpattern
ggplot(df, aes(x = category, y = value, fill = category)) +
  geom_col() +
  scale_fill_pattern(
    patterns = c("stripe", "stripe45", "grid", "brick"),
    fg = "black" # Foreground color for pattern lines
  ) +
  theme_minimal() +
  labs(
    title = "fillpattern Example",
    subtitle = "Simpler API with pattern names: stripe, grid, brick, etc."
  )
```

## 6.3 Comparing ggpattern vs fillpattern

```{r comparison-table, echo=FALSE}
comparison <- data.frame(
  Feature = c(
    "Pattern types",
    "Rendering performance",
    "Customization level",
    "Legend control",
    "Learning curve",
    "Best use case"
  ),
  ggpattern = c(
    "20+ types (stripe, circle, polygon, gradient, etc.)",
    "Slow (grid-based, can be very slow)",
    "High (many parameters: density, spacing, size, angle, etc.)",
    "Excellent (pattern_key_scale_factor)",
    "Moderate (many parameters to learn)",
    "Publication graphics, detailed customization"
  ),
  fillpattern = c(
    "12+ tile patterns (brick, stripe, grid, herringbone, hexagon, fish, etc.)",
    "Fast (optimized grid rendering)",
    "Moderate (angle, width/height, lwd, lty modifiers)",
    "Works with standard theme settings",
    "Easy (pattern name strings with modifiers)",
    "Fast rendering, tile-based patterns, production plots"
  )
)

knitr::kable(comparison,
  caption = "ggpattern vs fillpattern Comparison",
  col.names = c("Feature", "ggpattern", "fillpattern")
)
```

---

# Part 7: Best Practices and Tips

## 7.1 Essential Guidelines

### Always Set pattern_key_scale_factor

```r
# DO THIS
geom_*_pattern(..., pattern_key_scale_factor = 1.5)

# NOT THIS (patterns invisible in legend)
geom_*_pattern(...)
```

### Choose Appropriate Patterns

- **For print (B&W)**: Use stripe, crosshatch, circle
- **For color**: Combine patterns with distinct colors
- **For accessibility**: Always use patterns + colors, never color alone

### Optimize Performance

```r
# Slow (avoid for large data)
pattern = "plasma"
pattern = "gradient"

# Fast (good for large data)
pattern = "stripe"
pattern = "circle"
pattern = "none"  # Use when pattern not needed
```

## 7.2 Common Patterns Combinations

| Pattern 1 | Pattern 2 | Pattern 3 | Use Case |
|-----------|-----------|-----------|----------|
| stripe | crosshatch | circle | Classic trio, highly distinguishable |
| stripe (0°) | stripe (45°) | stripe (90°) | Same pattern, different angles |
| circle | regular_polygon | rose | All geometric shapes |
| stripe | none | crosshatch | Include "no pattern" option |

## 7.3 Troubleshooting

### Patterns Not Visible in Legend
→ Add `pattern_key_scale_factor = 1.5-2.0`

### Patterns Too Dense/Sparse
→ Adjust `pattern_spacing` (lower = denser) or `pattern_density`

### Slow Rendering
→ Switch from gradient/plasma to stripe/circle patterns
→ Consider using fillpattern for large datasets

### Patterns Don't Match Aesthetic
→ Check `scale_pattern_manual()` values match your pattern types
→ Ensure factor levels match pattern assignments

---

# Key Takeaways

## Essential Pattern Parameters Summary

| Parameter | What It Does | Effect When Increased | Effect When Decreased | Recommended Values |
|-----------|--------------|----------------------|----------------------|-------------------|
| **`pattern`** | Sets the pattern type | N/A | N/A | "stripe", "crosshatch", "circle" |
| **`pattern_density`** | Controls number of pattern elements per unit area | More elements appear, pattern looks busier/denser | Fewer elements, pattern looks sparser | 0.3-0.7 for most uses |
| **`pattern_spacing`** | Distance between pattern elements | Elements spread apart, pattern becomes sparser (**inverse relationship**) | Elements pack together, pattern becomes denser | 0.02-0.03 for stripes |
| **`pattern_angle`** | Rotation of pattern in degrees | Pattern rotates clockwise | Pattern rotates counter-clockwise | 0 (vertical), 45 (diagonal), 90 (horizontal) |
| **`pattern_size`** | Size of individual pattern elements (circles, polygons) | Shapes become larger, fewer fit in space | Shapes become smaller, more fit in space | 8-15 for circles |
| **`pattern_fill`** | Color of the pattern elements themselves | N/A | N/A | Contrasting color to base fill |
| **`pattern_colour`** | Outline/stroke color around pattern elements | N/A | N/A | Usually "black" or same as pattern_fill |
| **`pattern_key_scale_factor`** | Scales pattern density in legend keys only | Patterns more visible in legend | Patterns less visible in legend | **1.5-2.0 (always set this!)** |
| **`pattern_shape`** | Number of sides for regular_polygon | More sides (approaches circle) | Fewer sides (triangle at n=3) | 3-8 for clarity |

## Pattern Categories

1. **Stripe patterns**: stripe, crosshatch, wave, weave
   - Most commonly used and relatively fast
   - Control with: `pattern_angle`, `pattern_spacing`, `pattern_density`

2. **Geometric patterns**: circle, regular_polygon, rose, pch
   - Good for distinct visual encoding
   - Control with: `pattern_size`, `pattern_spacing`, `pattern_shape`

3. **Noise/gradient patterns**: gradient, plasma, ambient
   - Visually appealing but **very slow to render**
   - Control with: `pattern_fill`, `pattern_fill2`, `pattern_frequency`

## Critical Functions

| Function | Purpose | Example |
|----------|---------|---------|
| `geom_*_pattern()` | Pattern versions of standard geoms | `geom_col_pattern()`, `geom_boxplot_pattern()` |
| `scale_pattern_manual()` | Manually assign pattern types to levels | `scale_pattern_manual(values = c("stripe", "circle"))` |
| `scale_pattern_density_*()` | Control density scales | `scale_pattern_density_continuous(range = c(0.1, 0.9))` |
| `scale_pattern_fill_*()` | Control pattern fill colors | `scale_pattern_fill_manual(values = c("red", "blue"))` |
| `pattern_key_scale_factor` | Make patterns visible in legends (**most important!**) | `pattern_key_scale_factor = 1.8` |

## Best Practices Checklist

✅ **Always** set `pattern_key_scale_factor = 1.5-2.0`

✅ Combine `theme(legend.key.size = unit(1.5, "cm"))` with pattern scaling

✅ Use `pattern_spacing` between 0.02-0.04 for readable stripe patterns

✅ For accessibility, use patterns + colors together, never rely on color alone

✅ Test patterns in B&W to ensure they're distinguishable

✅ Keep it simple: don't map too many pattern aesthetics at once (max 2-3)

✅ For large datasets or many panels, consider fillpattern instead

⚠️ **Performance**: ggpattern can be very slow, especially with gradient/plasma patterns

❌ **Avoid**: Using gradient/plasma patterns on large datasets (extremely slow)

---

# Session Info

```{r session-info}
sessionInfo()
```
