---
title: "R graphics"
author: "Zhenguo Zhang"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
  github_document:
    toc: true
  html_document:
    toc: true
---

## Fonts

### Functions

Function | Example | Description
--- | --- | ---
quantzFonts | quartzFonts(avenir = c(“Avenir Book”, “Avenir Black”, “Avenir Book Oblique”, “Avenir Black Oblique”)) | convert system fonts to quantz fonts for use in R as families.


### Packages

- extrafont: import system fonts into R, https://r-coder.com/custom-fonts-r/
- ggVennDiagram: plot venn diagram based on the ggplot2

## Element size

### Functions

Function | Example | Description
--- | --- | ---
strwidth | strwidth("how wide", units="inches", cex=1, font=1, family="sans") | get the width of a string in current plotting device. `units` can be 'figure' so that the coordinates are normalized to the [0,1] range.
strheight | strheight("how high", units="inches", cex=1, font=1, family="sans") | get the height of a string in current plotting device.

### Discussion of *strwidth()* and *strheight()*

The two functions *strwidth()* and *strheight()* can measure the
width and height of a string. In default, the used font size is
9.75. So to get the size for other font size, one need to use the
parameter *cex*.

Also, the functions have 3 choices for the parameter *units*, which
have the following meaning:

* user: the default, using the plot's coordinate system to give the
  width and height, only available if plot.new() has been called.
  
* inches: give the width and height using inch as units.

* figure: give the width and height as fractions of figure size, so
  between 0 and 1.
  

As an example, we show the size of the letter **A** in the following
plot.

```{r}
plot(NULL,xlim=c(0,10), ylim=c(0,10))
aw<-strwidth("A")
ah<-strheight("A")
# draw the character to show the width
text(2,5,"A", pos=4, offset=0)
abline(v=c(2,2+aw), col="red", lty="dashed")
# similarly, show the height
text(5,2,"A",pos=3,offset=0)
abline(h=c(2,2+ah), col="blue", lty="dashed")
# let's try a letter with cex=2
aw<-strwidth("A", cex=2)
ah<-strheight("A", cex=2)
text(4,8,"A", pos=4, offset=0, cex=2)
abline(v=c(4,4+aw), col="red", lty="dashed")
text(8,4,"A", pos=3, offset=0, cex=2)
abline(h=c(4,4+ah), col="red", lty="dashed")
```

One can convert the measurements from the 3 different units. Before
we start, we need to introduce several important graphic parameters:

* fin: a vector of (width, height), which provides the figure size
  in inches. These values depend on the graphic device being used,
  so if one expands/shrinks the device, the values will change in the
  same direction.
  
* plt: A vector of the form c(x1, x2, y1, y2) giving the coordinates
  of the plot region as fractions of the current figure region, so these
  values are between 0 and 1.

* pin: a vector of (width, height), which provides the plot region
  size in inches.
  
* usr: a vector of the form c(x1, x2, y1, y2) giving the extremes of
  the user coordinates of the plotting region, i.e., the values at
  the ends of axes.
  
One can see `par()` for more information on these parameters. In the following
code, we will show how to convert the values of different units, using
width as an example.
  
```{r}
plot(NULL,xlim=c(0,10), ylim=c(0,10))
aw<-strwidth("A")
aw.i<-strwidth("A", units="inch")
aw.f<-strwidth("A", units="figure")
fin<-par("fin")
# conversion between inch and figure
all.equal(aw.i/fin[1], aw.f)
# convert between user and inch
xylims<-par("usr")
pltSize<-par("pin")
xrange<-xylims[2]-xylims[1]
all.equal(aw/xrange*pltSize[1], aw.i)
# convert between user and figure
all.equal(aw/xrange*pltSize[1]/fin[1], aw.f)
```

## Combine and arrange plots

packages: 

- patchwork: to reduce the space between plots, one can add an empty space via plot_spacer() and then
    set the spacer's width/height to a negative value in plot_layout().
    To modify the themes of all the plots, one can use the operator '&' instead of '+' (which only
    modifies the last plot), and the operator '*' only affects subplots in current nesting level.
- egg, 
- cowplot: this package separates the alignment and arrangement steps, making it more flexible in aligning
    different elements of plots, such as axis, titles, plots, etc; otherwise, it has similar functions as
    patchwork. To adjust space between plots, one can add an empty plot in between, and then adjust the
    options `rel_widths` and `rel_heights` in the function plot_grid(); the values can be set to negative
    to shrink the spaces.
- plotly::subplot(): used to combine plotly objects. examples are here https://plotly-r.com/arranging-views.html.
    To adjust the spaces between subplots, one can use the parameter `margin`, which will be applied to
    each subplot.
- ggiraph: function girafe() can be used to combine plots. It essentially combines the ggplot objects.


## Interactive plot

packages: 

- plotly, 
- ggiraph: it can convert nearly all ggplot elements (geoms, legends, scales, guides, etc) into interactive.
    see [here](https://davidgohel.github.io/ggiraph/articles/offcran/using_ggiraph.html) for examples.
- autoplotly: generate interactive plots on results supported by ggfortify, implemented with plotly.js and ggplot2.


## Figure format

### Converter packages

ggplotify: convert different graphics into grobs.


