---
title: "mcptools: Model Context Protocol for R"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

`mcptools` is an R package that implements the **Model Context Protocol (MCP)**, an open-source standard for connecting AI applications to external systems. Think of MCP like a USB-C port for AI applications—it provides a standardized way to connect AI tools to data sources, tools, and workflows.

### What is the Model Context Protocol?

MCP enables AI applications like Claude or ChatGPT to connect to:
- **Data sources**: local files, databases, APIs
- **Tools**: search engines, calculators, R functions
- **Workflows**: specialized prompts, custom logic

### What can mcptools do?

`mcptools` enables R to work with MCP in two ways:

1. **R as an MCP Server**: AI coding assistants (Claude Desktop, VS Code Copilot, Claude Code) can run R code in your active R sessions to answer questions about your data and environment
2. **R as an MCP Client**: Register third-party MCP servers to add external capabilities to R-based chat applications (e.g., `shinychat`, `querychat`)

### Use Cases

- AI assistants can inspect your R environment, check package documentation, and analyze your data
- Claude Code can generate data visualizations by accessing your actual datasets
- Enterprise chatbots can query R-connected databases and run statistical analyses
- Shiny chat applications can integrate external tools like web search or file systems

## Installation & Setup

### Installation

```{r install}
# From CRAN (stable version)
install.packages("mcptools")

# Development version from GitHub
# install.packages("pak")
pak::pak("posit-dev/mcptools")
```

### Verify Installation

```{r verify}
library(mcptools)
packageVersion("mcptools")
```

### Prerequisites

- R 4.1.0 or higher
- For server mode: An MCP-compatible AI client (Claude Desktop, Claude Code, VS Code with Copilot)
- For client mode: The `ellmer` package for chat functionality

## Key Concepts

| Concept | Description | Example |
|---------|-------------|---------|
| **MCP Server** | A system that exposes tools/data to AI clients | R session exposing functions to Claude Desktop |
| **MCP Client** | An AI application that connects to servers | Claude Desktop, ChatGPT, R chat apps using `ellmer` |
| **Tools** | Functions that AI clients can call | `rnorm()`, `read.csv()`, database queries |
| **Transport** | Communication method between client and server | `stdio` (standard input/output) or `http` |
| **Session** | An interactive R session exposed via MCP | Your running R console with loaded data/packages |

### Architecture Overview

**R as MCP Server:**
```
[AI Client: Claude Desktop] <--MCP--> [R Session: mcptools server] <--> [Your R Environment]
```

**R as MCP Client:**
```
[R Chat App: shinychat] <---> [mcptools client] <--MCP--> [External MCP Servers]
```

## Key Functions/Commands

| Function | Purpose | Example | Key Parameters |
|----------|---------|---------|----------------|
| `mcp_server()` | Start an MCP server to expose R functions to AI clients | `mcp_server(tools = list(tool_rnorm))` | `tools`, `type`, `host`, `port`, `session_tools` |
| `mcp_tools()` | Fetch tools from MCP servers for use in R chat apps | `tools <- mcp_tools(config = "~/.config/mcp.json")` | `config` (path to config file) |
| `mcp_session()` | Make an interactive R session available to MCP servers | `mcp_session()` | None |
| `tool()` | Define a custom tool (from `ellmer` package) | `tool(rnorm, description = "Generate random numbers", arguments = list(...))` | `fun`, `description`, `arguments`, `name` |

### Function Details

#### `mcp_server()`

Starts an MCP server that exposes R functions as tools.

```r
mcp_server(
  tools = NULL,
  ...,
  type = c("stdio", "http"),
  host = "127.0.0.1",
  port = as.integer(Sys.getenv("MCPTOOLS_PORT", "8080")),
  session_tools = TRUE
)
```

**Parameters:**
- `tools`: List of ellmer tools to expose, or path to .R file with tool definitions (default: `NULL`)
- `...`: Reserved for future use
- `type`: Transport type - `"stdio"` (default, for local) or `"http"` (for network)
- `host`: Host to bind for HTTP transport (default: `"127.0.0.1"`)
- `port`: Port for HTTP transport (default: from `MCPTOOLS_PORT` env var or `8080`)
- `session_tools`: Include built-in session tools (default: `TRUE`)

**Returns:** Blocks R process and runs server until stopped

#### `mcp_tools()`

Fetches tools from configured MCP servers for use in chat applications.

```r
mcp_tools(config = NULL)
```

**Parameters:**
- `config`: Path to MCP servers configuration file (JSON format). Default: `.mcptools_config` option or `~/.config/mcptools/config.json`

**Returns:** List of ellmer tools compatible with `Chat` objects

#### `mcp_session()`

Makes the current R session available to MCP servers, allowing AI clients to interact with your environment.

```r
mcp_session()
```

**Parameters:** None

**Returns:** Invisibly returns nanonext socket for communication

#### `tool()` (from ellmer package)

Define a custom tool for use with MCP or chat applications.

```r
tool(
  fun,
  description,
  arguments = list(),
  name = NULL,
  convert = TRUE,
  annotations = list()
)
```

**Parameters:**
- `fun`: The R function to be invoked when the tool is called
- `description`: Detailed description of what the function does
- `arguments`: Named list of argument definitions using `type_*()` functions
- `name`: Tool name (optional if `fun` is a named function)
- `convert`: Auto-convert JSON inputs to R types (default: `TRUE`)
- `annotations`: Additional metadata (usually from `tool_annotations()`)

## Configuration

### Configuring Claude Desktop to Use R as Server

To allow Claude Desktop to run R code in your sessions, add this to your Claude Desktop configuration file:

**Location:**
- macOS: `~/Library/Application Support/Claude/claude_desktop_config.json`
- Windows: `%APPDATA%\Claude\claude_desktop_config.json`
- Linux: `~/.config/Claude/claude_desktop_config.json`

**Basic Configuration:**
```json
{
  "mcpServers": {
    "r-mcptools": {
      "command": "Rscript",
      "args": ["-e", "mcptools::mcp_server()"]
    }
  }
}
```

**With Custom Tools:**
```json
{
  "mcpServers": {
    "r-custom": {
      "command": "Rscript",
      "args": ["-e", "mcptools::mcp_server(tools = '/path/to/tools.R')"]
    }
  }
}
```

### Configuring Claude Code

Use the command line to add R as an MCP server:

```bash
claude mcp add -s "user" r-mcptools -- Rscript -e "mcptools::mcp_server()"
```

Or manually edit: `~/.config/Claude Code/mcp_config.json`

### Configuration File for MCP Tools (Client Mode)

When using R as an MCP client, create a configuration file listing the servers to connect to:

**Default location:** `~/.config/mcptools/config.json`

```json
{
  "mcpServers": {
    "filesystem": {
      "command": "npx",
      "args": ["-y", "@modelcontextprotocol/server-filesystem", "/home/user/data"]
    },
    "github": {
      "command": "npx",
      "args": ["-y", "@modelcontextprotocol/server-github"]
    }
  }
}
```

## Full Working Examples

### Example 1: R as an MCP Server - Custom Random Number Tool

Create a custom tool that generates random numbers and expose it to AI clients:

```{r server-example}
library(mcptools)
library(ellmer)

# Define a custom tool for generating random numbers
# Note: arguments are now passed as a named list
tool_rnorm <- tool(
  rnorm,
  description = "Draw numbers from a random normal distribution",
  arguments = list(
    n = type_integer("Number of observations to generate"),
    mean = type_number("Mean value of the distribution"),
    sd = type_number("Standard deviation of the distribution")
  )
)

# Define another tool for summary statistics
tool_summary <- tool(
  summary,
  description = "Generate summary statistics for a numeric vector",
  arguments = list(
    object = type_object("The object to summarize")
  )
)

# Start the MCP server with these tools
# This will block and wait for connections from AI clients
mcp_server(tools = list(tool_rnorm, tool_summary))
```

**Usage:** After starting this server and configuring Claude Desktop, you can ask:
- "Generate 100 random numbers with mean 50 and sd 10"
- "Create a normal distribution and show me its summary"

### Example 2: R as an MCP Server - Data Analysis Tool

Create a tool that performs data analysis on your loaded datasets:

```{r analysis-server}
library(mcptools)
library(ellmer)

# Define a tool for data exploration
tool_explore_data <- tool(
  function(dataset_name) {
    # Get the dataset from the global environment
    data <- get(dataset_name, envir = .GlobalEnv)

    list(
      dimensions = dim(data),
      column_names = names(data),
      column_types = sapply(data, class),
      summary = summary(data),
      head = head(data, 3)
    )
  },
  name = "explore_data",
  description = "Explore a dataset in the R environment",
  arguments = list(
    dataset_name = type_string("Name of the dataset to explore")
  )
)

# Load some example data
data(mtcars)
data(iris)

# Start the server
mcp_server(tools = list(tool_explore_data))
```

**Usage:** Ask Claude Desktop: "Explore the mtcars dataset and tell me about it"

### Example 3: R as an MCP Client - Chat App with External Tools

Use mcptools in an ellmer chat application to access external MCP servers:

```{r client-example}
library(mcptools)
library(ellmer)

# First, create a config file at ~/.config/mcptools/config.json
# with your desired MCP servers (filesystem, GitHub, etc.)

# Load tools from MCP servers
external_tools <- mcp_tools(config = "~/.config/mcptools/config.json")

# Create a chat object
chat <- chat_anthropic()

# Set the tools from MCP servers
chat$set_tools(external_tools)

# Now you can chat with access to external tools
chat$chat("What issues are open on posit-dev/mcptools?")
```

**Note:** This assumes you have configured a GitHub MCP server in your config file.

### Example 4: Registering an Interactive R Session

Allow AI assistants to interact with your running R session:

```{r session-example}
library(mcptools)

# Load some data and packages you want to work with
library(dplyr)
library(ggplot2)
data(mtcars)

# Register this session with MCP
# Add this to your .Rprofile to automatically register new sessions
mcp_session()

# Now AI assistants can:
# - See what objects are in your environment
# - Check which packages are loaded
# - Access help for functions in loaded packages
# - Suggest analyses based on your data
```

**Usage:** Once configured in Claude Desktop, ask:
- "What datasets do I have loaded?"
- "Show me the structure of mtcars"
- "What does the filter function in dplyr do?"

### Example 5: Using the btw Package Integration

The `btw` package provides excellent default tools for R sessions:

```{r btw-integration}
# Install btw package
# install.packages("btw")

library(mcptools)
library(btw)

# The btw package automatically provides tools when loaded
# Start a server with btw's built-in tools:
# - Browse package documentation
# - Inspect objects in your environment
# - Get session info

mcp_server()  # btw tools are included automatically if btw is loaded
```

**Usage with btw:**
- "What functions are available in the dplyr package?"
- "Show me the structure of the iris dataset in my environment"
- "What packages do I have loaded?"

### Example 6: HTTP Server for Remote Access

Run an MCP server over HTTP for network access:

```{r http-server}
library(mcptools)
library(ellmer)

# Define tools
tool_system_info <- tool(
  function() {
    list(
      r_version = R.version.string,
      platform = Sys.info()["sysname"],
      working_dir = getwd()
    )
  },
  name = "system_info",
  description = "Get R system information",
  arguments = list()  # No arguments
)

# Start HTTP server
mcp_server(
  tools = list(tool_system_info),
  type = "http",
  host = "0.0.0.0",  # Listen on all interfaces
  port = 8080
)

# Access from another machine via http://your-ip:8080
```

**Note:** Be cautious with HTTP servers - only expose them on trusted networks.

### Example 7: Tool with Error Handling

Create a safe tool with proper error handling:

```{r error-tool}
library(mcptools)
library(ellmer)

tool_safe_divide <- tool(
  function(x, y) {
    tryCatch({
      if (y == 0) {
        return(list(error = TRUE, message = "Cannot divide by zero"))
      }
      list(result = x / y, error = FALSE)
    }, error = function(e) {
      list(error = TRUE, message = conditionMessage(e))
    })
  },
  name = "safe_divide",
  description = "Safely divide two numbers with error handling",
  arguments = list(
    x = type_number("Numerator"),
    y = type_number("Denominator")
  )
)

mcp_server(tools = list(tool_safe_divide))
```

## Common Patterns & Best Practices

### 1. Tool Design Patterns

**Keep tools focused on single responsibilities:**
```{r tool-pattern-1}
library(ellmer)

# Good: Single responsibility
tool_mean <- tool(
  mean,
  description = "Calculate the arithmetic mean of a numeric vector",
  arguments = list(
    x = type_number("Numeric vector")
  )
)

# Avoid: Multiple responsibilities in one tool
# Better to create separate tools for mean, median, sd, etc.
```

**Provide detailed descriptions:**
```{r tool-pattern-2}
# Good: Detailed description helps the LLM understand when to use it
tool_filter_data <- tool(
  function(data_name, condition) {
    data <- get(data_name, envir = .GlobalEnv)
    subset(data, eval(parse(text = condition)))
  },
  name = "filter_data",
  description = "Filter rows of a data frame based on a logical condition. The condition should be a string representing valid R code that evaluates to a logical vector. For example: 'mpg > 20' or 'cyl == 6'.",
  arguments = list(
    data_name = type_string("Name of the data frame in the global environment"),
    condition = type_string("Logical condition as a string (e.g., 'mpg > 20')")
  )
)
```

### 2. Use Type Specifications Correctly

```{r type-specs}
library(ellmer)

# Available type functions:
# - type_string()    : For text
# - type_number()    : For numeric values
# - type_integer()   : For whole numbers
# - type_boolean()   : For TRUE/FALSE
# - type_object()    : For complex objects
# - type_array()     : For arrays/lists
# - type_enum()      : For specific choices

tool_example <- tool(
  function(text, count, enabled, choice) {
    # Function implementation
  },
  description = "Example showing different type specifications",
  arguments = list(
    text = type_string("A text parameter"),
    count = type_integer("A whole number"),
    enabled = type_boolean("A true/false flag"),
    choice = type_enum(c("option1", "option2", "option3"),
                       description = "Pick one option")
  )
)
```

### 3. Session Management

**Run server in background for production:**
```{r session-management}
# Option 1: Run in separate R process
# Start a new R session and run:
library(mcptools)
mcp_server(tools = my_tools)

# Option 2: Use callr for background process
library(callr)
r_bg <- r_bg(function() {
  mcptools::mcp_server()
})

# Check if it's running
r_bg$is_alive()

# Stop it when done
r_bg$kill()
```

### 4. Security Considerations

**Validate inputs and restrict access:**
```{r security-example}
library(ellmer)

# Bad: Unrestricted file access
tool_read_file_unsafe <- tool(
  readLines,
  description = "Read any file",
  arguments = list(
    con = type_string("File path")
  )
)

# Good: Restricted to specific directory
tool_read_data_safe <- tool(
  function(filename) {
    # Restrict to safe directory
    safe_dir <- "~/safe_data_dir"
    safe_path <- file.path(safe_dir, basename(filename))

    # Validate the path is within safe directory
    if (!file.exists(safe_path)) {
      stop("File not found in safe directory")
    }

    readLines(safe_path)
  },
  name = "read_safe_data",
  description = "Read file from safe data directory only",
  arguments = list(
    filename = type_string("Filename (basename only, no paths)")
  )
)
```

**Security best practices:**
- Never expose sensitive functions over HTTP without authentication
- Validate all inputs in custom tools
- Use `stdio` transport for local-only access
- Be cautious with file system and database access
- Review tool code before exposing to AI clients

### 5. Testing Your Tools

Before deploying MCP servers, test tools locally:

```{r testing}
library(mcptools)
library(ellmer)

# Define tool
my_tool <- tool(
  function(x) x^2,
  name = "square",
  description = "Square a number",
  arguments = list(
    x = type_number("Number to square")
  )
)

# Test the underlying function directly
my_tool$fun(5)  # Should return 25

# Test with a chat (requires API key)
chat <- chat_anthropic()
chat$register_tool(my_tool)
chat$chat("What is 7 squared?")

# Check the conversation history
chat
```

### 6. Working with R Objects

**Serialize complex R objects properly:**
```{r objects}
library(ellmer)

tool_get_model_summary <- tool(
  function(model_name) {
    model <- get(model_name, envir = .GlobalEnv)

    # Return a list that can be serialized to JSON
    list(
      coefficients = coef(model),
      r_squared = summary(model)$r.squared,
      residual_se = summary(model)$sigma,
      n_obs = nobs(model)
    )
  },
  name = "get_model_summary",
  description = "Get summary statistics from a fitted linear model",
  arguments = list(
    model_name = type_string("Name of the model object in environment")
  )
)
```

## Common Pitfalls to Avoid

1. **Blocking the R session**: `mcp_server()` blocks the current R session. Run it in a separate process if you need your session for other work.

2. **Forgetting to restart AI clients**: After changing MCP configuration, restart Claude Desktop or VS Code to pick up changes.

3. **Incorrect tool parameter format**: In ellmer 0.3.0+, use `arguments = list(...)` not `...` directly.

4. **Path issues**: Use absolute paths in configuration files, especially on Windows.

5. **Port conflicts**: If using HTTP transport, ensure the port isn't already in use with `netstat -an | grep <port>`.

6. **Missing type specifications**: Always define parameter types with `type_*()` functions for proper tool schemas.

7. **Not testing tools locally**: Test tool functions independently before exposing them via MCP.

## Troubleshooting

### Server doesn't start
- Check that `Rscript` is in your PATH: `which Rscript` (Unix) or `where Rscript` (Windows)
- Verify the configuration file syntax is valid JSON
- Look for error messages in the AI client's developer logs
- Try running `Rscript -e "mcptools::mcp_server()"` directly in terminal

### Tools not appearing in AI client
- Ensure `session_tools = TRUE` if you want built-in tools
- Verify tool definitions use correct `tool()` syntax (ellmer 0.3.0+)
- Restart the AI client application completely
- Check that the R process is still running

### Connection issues with HTTP server
- Check firewall settings: `sudo ufw status` (Linux)
- Verify the correct host and port in configuration
- Ensure the server process is still running: `ps aux | grep Rscript`
- Test connection: `curl http://localhost:8080` or `nc -zv localhost 8080`

### Tools failing silently
- Add comprehensive error handling to tool functions
- Check R's warning/error messages
- Test tools locally first with `tool_name$fun(args)`
- Enable debug logging in the AI client

### Chat not using tools
- Verify tools are registered: `chat$get_tools()`
- Ensure tool descriptions are clear and detailed
- Try asking more explicitly: "Use the X tool to do Y"
- Check that tool parameter types match what you're asking for

## References & Further Reading

### Official Documentation
- [mcptools GitHub Repository](https://github.com/posit-dev/mcptools) - Source code and latest development
- [mcptools CRAN Page](https://cran.r-project.org/package=mcptools) - Stable release documentation
- [mcptools Package Site](https://posit-dev.github.io/mcptools/) - Full package documentation
- [Model Context Protocol Official Site](https://modelcontextprotocol.io) - MCP specification and overview
- [MCP Documentation](https://modelcontextprotocol.io/docs) - Complete MCP guides
- [MCP Architecture](https://modelcontextprotocol.io/docs/learn/architecture) - Protocol architecture details

### Related R Packages
- [ellmer](https://ellmer.tidyverse.org/) - R package for building LLM-powered applications
  - [Tool Calling Vignette](https://ellmer.tidyverse.org/articles/tool-calling.html)
- [btw](https://github.com/posit-dev/btw) - Contextual R help for LLMs, default MCP tools
- [shinychat](https://github.com/posit-dev/shinychat) - Shiny chat interface components
- [querychat](https://github.com/posit-dev/querychat) - Query-based chat applications

### MCP Ecosystem
- [MCP Registry](https://modelcontextprotocol.io/registry) - Directory of available MCP servers
- [MCP Specification](https://spec.modelcontextprotocol.io/) - Technical protocol specification
- [MCP GitHub](https://github.com/modelcontextprotocol) - Official MCP repositories

### AI Clients Supporting MCP
- [Claude Desktop](https://claude.ai/download) - Desktop app with MCP support
- [Claude Code](https://docs.anthropic.com/en/docs/claude-code) - AI coding assistant
- [VS Code with GitHub Copilot](https://code.visualstudio.com/docs/copilot/chat/mcp-servers) - Editor integration
- [Cursor](https://cursor.com/docs/context/mcp) - AI-powered code editor
- [ChatGPT Desktop](https://help.openai.com/en/articles/10204709-mcp) - OpenAI's MCP client

### Community Resources
- [Posit Community](https://community.rstudio.com/) - R community forum
- [MCP GitHub Discussions](https://github.com/modelcontextprotocol/specification/discussions) - MCP community

### Example MCP Servers (not R-specific)
- [@modelcontextprotocol/server-filesystem](https://github.com/modelcontextprotocol/servers/tree/main/src/filesystem) - File system access
- [@modelcontextprotocol/server-github](https://github.com/modelcontextprotocol/servers/tree/main/src/github) - GitHub API
- [@modelcontextprotocol/server-postgres](https://github.com/modelcontextprotocol/servers/tree/main/src/postgres) - PostgreSQL access
- [MCP Servers Repository](https://github.com/modelcontextprotocol/servers) - Official server implementations
