---
title: "plumber2: Building REST APIs in R"
output:
  github_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
  md_document:
    variant: gfm
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE, # Don't run API servers during knitting
  message = FALSE,
  warning = FALSE
)
```

# Introduction

**plumber2** is a modern R package for creating REST APIs that expose R functions as web services. It's a complete rewrite of the original plumber package with improved performance, better design patterns, and enhanced functionality.

# Why plumber2?

## plumber2 vs Shiny

While both plumber2 and Shiny allow you to expose R functionality over HTTP, they serve different purposes:

| Feature | plumber2 | Shiny |
|---------|----------|-------|
| **Purpose** | REST API / Web Services | Interactive Web Applications |
| **Interface** | JSON/CSV/XML endpoints | HTML/JavaScript UI |
| **Interaction** | Programmatic (via HTTP) | Visual/Interactive |
| **State** | Stateless (each request independent) | Stateful (maintains session) |
| **Use Case** | Data services, microservices, backend APIs | Dashboards, interactive visualizations |
| **Clients** | Any HTTP client (curl, Python, JavaScript) | Web browsers |
| **Output** | Structured data (JSON, CSV, etc.) | Web pages with widgets |

**When to use plumber2:**
- Building APIs for mobile apps or web applications
- Creating microservices that other systems consume
- Exposing R models/analysis as web services
- Need programmatic access to R functions
- Building backend services for data pipelines

**When to use Shiny:**
- Creating interactive dashboards for end users
- Need visual exploration tools
- Users interact through browser UI
- Require reactive programming patterns

# Creating APIs

## Basic API Structure

plumber2 uses a functional, pipe-friendly approach:

```{r basic-api}
library(plumber2)

# Create an API object
my_api <- api()

# Add endpoints using pipes
my_api |>
  api_get("/hello", function(name = "World") {
    list(greeting = paste("Hello,", name))
  })
```

## Annotation-Based Approach

You can also define APIs using special comments (`#*`) in `.R` files:

```{r annotation-file, eval=FALSE}
# plumber.R file
#* Echo back the input
#* @get /echo
function(msg = "Hello") {
  list(message = msg)
}

#* Calculate sum
#* @post /sum
function(x, y) {
  list(result = as.numeric(x) + as.numeric(y))
}
```

Load and run annotation-based APIs:

```{r load-annotations}
# Parse the file
api <- api("plumber.R")

# Run the server
api |> api_run(port = 8080)
```

## Key API Functions

| Function | Description |
|----------|-------------|
| `api()` | Create a new API object |
| `api_get()` | Add GET endpoint |
| `api_post()` | Add POST endpoint |
| `api_put()` | Add PUT endpoint |
| `api_delete()` | Add DELETE endpoint |
| `api_patch()` | Add PATCH endpoint |
| `api_run()` | Start the API server |
| `parse_plumber_file()` | Load annotation-based API from file |

# Input Types and Methods

## GET Requests

GET requests receive data through query parameters:

```{r get-request}
library(plumber2)

api() |>
  api_get("/search", function(query, limit = 10) {
    list(
      query = query,
      limit = as.integer(limit),
      results = paste("Searching for:", query)
    )
  })

# Access: http://localhost:8080/search?query=data&limit=20
```

## POST Requests

POST requests typically receive data in the request body:

```{r post-request}
api() |>
  api_post("/create", function(name, age, email) {
    list(
      status = "created",
      user = list(
        name = name,
        age = as.integer(age),
        email = email
      )
    )
  })

# Send JSON body: {"name": "Alice", "age": 30, "email": "alice@example.com"}
```

## Path Parameters

Extract variables from the URL path:

```{r path-params}
api() |>
  api_get("/users/{id}", function(id) {
    list(
      user_id = id,
      name = paste("User", id)
    )
  })

# Access: http://localhost:8080/users/123
```

## Request Object Access

Access the full request object for headers, cookies, body:

```{r request-object}
api() |>
  api_post("/upload", function(req) {
    list(
      content_type = req$CONTENT_TYPE,
      content_length = req$CONTENT_LENGTH,
      headers = names(req),
      body_size = length(req$postBody)
    )
  })
```

## Parameter Type Validation

Specify parameter types and validation:

```{r param-types}
api() |>
  api_get("/validate", function(
    age = numeric(), # Must be numeric
    name = character(), # Must be character
    active = logical() # Must be logical/boolean
  ) {
    list(
      age = age,
      name = name,
      active = active
    )
  })
```

# Handlers

## What are Handlers?

Handlers are R functions that execute when an incoming HTTP request matches their specified route. They form the core of your API's functionality, translating HTTP requests into R code execution.

**Key characteristics:**
- Standard R functions annotated with HTTP method decorators (`@get`, `@post`, etc.)
- Automatically receive parsed input from various sources (path, query, body, headers)
- Return values are automatically serialized to HTTP responses
- Can access special objects (`req`, `res`) for advanced control

## Handler Input Arguments

Handlers can receive several special arguments that provide access to request data:

| Argument | Type | Description |
|----------|------|-------------|
| `req` or `request` | Request object | Full HTTP request with headers, cookies, environment |
| `res` or `response` | Response object | Control response headers, status codes |
| Path parameters | Various | Dynamic values extracted from URL path using `<param>` syntax |
| `query` | List | Query string parameters (annotation-based) |
| Function parameters | Various | Mapped from query string (functional) or body (POST) |

**Important differences between functional and annotation approaches:**

- **Functional API (`api_get()`, `api_post()`, etc.)**:
  - Query parameters map directly to function arguments
  - Example: `function(name, age)` receives `?name=Alice&age=30`

- **Annotation-based API (`#* @get`, `#* @post`, etc.)**:
  - Path parameters use `<param>` syntax: `@get /users/<id>`
  - Query parameters require `query` object: `function(query)` then access `query$name`
  - Use `@query` annotations to document query parameters
  - Body parameters map directly to function arguments: `function(name, age)`

## Example 1: Basic Handler

### Functional Approach
```{r handler-basic-func}
library(plumber2)

api() |>
  api_get("/hello", function() {
    "hello world"
  })
```

### Annotation Approach
```{r handler-basic-annot}
#* Return hello world
#* @get /hello
function() {
  "hello world"
}
```

**How to test:**
```bash
curl http://localhost:8080/hello
# Output: ["hello world"]
```

## Example 2: Path Parameters

Extract dynamic values from the URL path using `<param>` syntax.

### Functional Approach
```{r handler-path-func}
api() |>
  api_get("/users/{id}", function(id) {
    list(
      user_id = id,
      name = paste("User", id)
    )
  })
```

### Annotation Approach
```{r handler-path-annot}
#* Lookup a user by ID
#* @get /users/<id>
function(id) {
  list(
    user_id = id,
    name = paste("User", id)
  )
}
```

**How to test:**
```bash
curl http://localhost:8080/users/123
# Output: {"user_id":["123"],"name":["User 123"]}
```

**Multiple path parameters:**
```{r handler-multi-path}
#* Calculate sum using path parameters
#* @get /sum/<x>/<y>
#* @param x The first number
#* @param y The second number
function(x, y) {
  list(result = as.numeric(x) + as.numeric(y))
}

# Test: curl http://localhost:8080/sum/5/10
# Output: {"result":[15]}
```

## Example 3: Request Object Access

Access headers, HTTP method, path, and other request metadata.

### Functional Approach
```{r handler-request-func}
api() |>
  api_get("/request-info", function(req) {
    list(
      method = req$REQUEST_METHOD,
      path = req$PATH_INFO,
      user_agent = req$HTTP_USER_AGENT,
      all_headers = names(req)
    )
  })
```

### Annotation Approach
```{r handler-request-annot}
#* Get request information
#* @get /request-info
function(req) {
  list(
    method = req$REQUEST_METHOD,
    path = req$PATH_INFO,
    user_agent = req$HTTP_USER_AGENT,
    all_headers = names(req)
  )
}
```

**How to test:**
```bash
curl http://localhost:8080/request-info
# Output: {"method":"GET","path":"/request-info","user_agent":"curl/...","all_headers":[...]}
```

## Example 4: Response Object (Custom Headers & Status)

Control HTTP status codes and response headers.

### Functional Approach
```{r handler-response-func}
api() |>
  api_get("/custom-response", function(res) {
    res$setHeader("X-Custom-Header", "MyValue")
    res$status <- 201
    list(message = "Created with custom headers")
  })
```

### Annotation Approach
```{r handler-response-annot}
#* Custom response with headers
#* @get /custom-response
function(res) {
  res$setHeader("X-Custom-Header", "MyValue")
  res$status <- 201
  list(message = "Created with custom headers")
}
```

**How to test:**
```bash
curl -v http://localhost:8080/custom-response
# Status: 201
# Header: X-Custom-Header: MyValue
# Body: {"message":"Created with custom headers"}
```

## Example 5: Query Parameters

Access URL query string parameters using the `query` object.

### Functional Approach
```{r handler-query-func}
api() |>
  api_get("/search", function(q, limit = "10", offset = "0") {
    list(
      search_term = q,
      limit = as.integer(limit),
      offset = as.integer(offset)
    )
  })
```

### Annotation Approach
```{r handler-query-annot}
#* Search with query parameters
#* @get /search
#* @query q Search term
#* @query limit Number of results (default 10)
#* @query offset Result offset (default 0)
function(query) {
  q <- query$q %||% ""
  limit <- query$limit %||% "10"
  offset <- query$offset %||% "0"

  list(
    search_term = q,
    limit = as.integer(limit),
    offset = as.integer(offset)
  )
}
```

**How to test:**
```bash
curl 'http://localhost:8080/search?q=test&limit=20&offset=5'
# Output: {"search_term":["test"],"limit":[20],"offset":[5]}
```

**Note:** In annotation-based APIs, query parameters come through the `query` object. Use `%||%` (or equivalent) to provide defaults.

## Example 6: POST with Body Parameters

Receive data from POST request body (JSON format).

### Functional Approach
```{r handler-post-func}
api() |>
  api_post("/users", function(name, email, age) {
    list(
      status = "created",
      user = list(
        name = name,
        email = email,
        age = as.integer(age)
      )
    )
  })
```

### Annotation Approach
```{r handler-post-annot}
#* Create a new user
#* @post /users
function(name, email, age) {
  list(
    status = "created",
    user = list(
      name = name,
      email = email,
      age = as.integer(age)
    )
  )
}
```

**How to test:**
```bash
curl -X POST http://localhost:8080/users \
  -H 'Content-Type: application/json' \
  -d '{"name":"Alice","email":"alice@example.com","age":30}'
# Output: {"status":"created","user":{"name":"Alice","email":"alice@example.com","age":30}}
```

## Example 7: Combined Inputs

Combine path parameters, query parameters, body, request, and response objects.

### Functional Approach
```{r handler-combined-func}
api() |>
  api_post("/items/{category}", function(category, priority = "1", name, price, req, res) {
    res$setHeader("X-Processed-By", "plumber2")

    list(
      category = category,
      priority = as.integer(priority),
      item = list(
        name = name,
        price = as.numeric(price)
      ),
      user_agent = req$HTTP_USER_AGENT,
      timestamp = as.character(Sys.time())
    )
  })
```

### Annotation Approach
```{r handler-combined-annot}
#* Combined inputs: path, query, body, req, res
#* @post /items/<category>
#* @query priority Priority level (default 1)
function(category, query, name, price, req, res) {
  priority <- query$priority %||% "1"
  res$setHeader("X-Processed-By", "plumber2")

  list(
    category = category,
    priority = as.integer(priority),
    item = list(
      name = name,
      price = as.numeric(price)
    ),
    user_agent = req$HTTP_USER_AGENT,
    timestamp = as.character(Sys.time())
  )
}
```

**How to test:**
```bash
curl -X POST 'http://localhost:8080/items/electronics?priority=5' \
  -H 'Content-Type: application/json' \
  -d '{"name":"Laptop","price":999}'
# Output: {"category":["electronics"],"priority":[5],"item":{"name":["Laptop"],"price":[999]},...}
```

## Example 8: Validation with Error Handling

Validate inputs and return errors when validation fails.

### Functional Approach
```{r handler-validation-func}
api() |>
  api_post("/validate", function(email, age) {
    if (!grepl("@", email)) {
      stop("Invalid email format")
    }

    age_num <- as.numeric(age)
    if (is.na(age_num) || age_num < 0 || age_num > 150) {
      stop("Invalid age value")
    }

    list(email = email, age = age_num, valid = TRUE)
  })
```

### Annotation Approach
```{r handler-validation-annot}
#* Validate email and age
#* @post /validate
function(email, age) {
  if (!grepl("@", email)) {
    stop("Invalid email format")
  }

  age_num <- as.numeric(age)
  if (is.na(age_num) || age_num < 0 || age_num > 150) {
    stop("Invalid age value")
  }

  list(email = email, age = age_num, valid = TRUE)
}
```

**How to test (valid):**
```bash
curl -X POST http://localhost:8080/validate \
  -H 'Content-Type: application/json' \
  -d '{"email":"test@example.com","age":25}'
# Output: {"email":"test@example.com","age":25,"valid":true}
```

**How to test (invalid):**
```bash
curl -X POST http://localhost:8080/validate \
  -H 'Content-Type: application/json' \
  -d '{"email":"invalid","age":25}'
# Output: {"error":"500 - Internal server error","message":"Invalid email format"}
```

## Example 9: Type Casting with Annotations

Specify parameter types for automatic validation and conversion. Use `@query` or `@param` with type annotations.

**Supported types:** `logical`, `boolean`, `numeric`, `double`, `integer`, `character`, `string`, `date`

### Annotation Approach (Path Parameters)
```{r handler-types-path}
#* Calculate sum with typed path parameters
#* @get /calc/<x>/<y>
#* @param x:numeric The first number
#* @param y:numeric The second number
function(x, y) {
  x_num <- as.numeric(x)
  y_num <- as.numeric(y)
  list(x = x_num, y = y_num, sum = x_num + y_num)
}

# Test: curl http://localhost:8080/calc/10.5/20.3
# Output: {"x":[10.5],"y":[20.3],"sum":[30.8]}
```

### Annotation Approach (Query Parameters)
```{r handler-types-query}
#* Calculate with typed query parameters
#* @get /calculate
#* @query x:numeric The first number
#* @query y:numeric The second number
#* @query round:logical Whether to round result
function(query) {
  x <- as.numeric(query$x %||% "0")
  y <- as.numeric(query$y %||% "0")
  round_it <- as.logical(query$round %||% "FALSE")

  result <- x + y
  if (round_it) {
    result <- round(result)
  }
  list(x = x, y = y, result = result, rounded = round_it)
}

# Test: curl 'http://localhost:8080/calculate?x=10.5&y=20.3&round=true'
# Output: {"x":[10.5],"y":[20.3],"result":[31],"rounded":[true]}
```

## Example 10: Required Parameters

Mark parameters as required using `*` after the type.

### Annotation Approach (Path Parameters)
```{r handler-required-path}
#* Safe calculator with required path parameters
#* @get /safe-calc/<a>/<b>
#* @param a:numeric* First number (required)
#* @param b:numeric* Second number (required)
function(a, b) {
  a_num <- as.numeric(a)
  b_num <- as.numeric(b)
  list(a = a_num, b = b_num, sum = a_num + b_num)
}

# Test: curl http://localhost:8080/safe-calc/10/5
# Output: {"a":[10],"b":[5],"sum":[15]}
```

### Annotation Approach (Query Parameters)
```{r handler-required-query}
#* Calculator with required query parameters
#* @get /calc-query
#* @query a:numeric* First number (required)
#* @query b:numeric* Second number (required)
#* @query operation:string Operation (add, subtract, multiply, divide)
function(query) {
  # Extract and validate required parameters
  if (is.null(query$a) || is.null(query$b)) {
    stop("Parameters 'a' and 'b' are required")
  }

  a <- as.numeric(query$a)
  b <- as.numeric(query$b)
  operation <- query$operation %||% "add"

  result <- switch(as.character(operation),
    "add" = a + b,
    "subtract" = a - b,
    "multiply" = a * b,
    "divide" = if (b != 0) a / b else NA,
    stop("Unknown operation")
  )
  list(a = a, b = b, operation = operation, result = result)
}

# Test addition:
# curl 'http://localhost:8080/calc-query?a=10&b=5&operation=add'
# Output: {"a":[10],"b":[5],"operation":["add"],"result":[15]}

# Test multiplication:
# curl 'http://localhost:8080/calc-query?a=10&b=5&operation=multiply'
# Output: {"a":[10],"b":[5],"operation":["multiply"],"result":[50]}
```

## Handler Best Practices

### 1. Use Appropriate HTTP Methods

Follow REST conventions:
- **GET**: Retrieve data (no side effects, idempotent)
- **POST**: Create new resources
- **PUT**: Update entire resources
- **PATCH**: Partial updates
- **DELETE**: Remove resources

### 2. Always Validate Inputs

```{r handler-best-validate}
api() |>
  api_post("/create", function(email, age) {
    # Validate email format
    if (!grepl("^[^@]+@[^@]+\\.[^@]+$", email)) {
      stop("Invalid email format")
    }

    # Validate age range
    age_num <- as.numeric(age)
    if (is.na(age_num) || age_num < 0 || age_num > 150) {
      stop("Age must be between 0 and 150")
    }

    list(email = email, age = age_num, status = "validated")
  })
```

### 3. Use Type Hints for Automatic Validation

```{r handler-best-types}
#* Use type annotations instead of manual casting
#* @get /calculate
#* @param x:numeric* Required number
#* @param y:numeric* Required number
function(x, y) {
  # x and y are already numeric
  list(sum = x + y, product = x * y)
}
```

### 4. Return Meaningful Error Responses

```{r handler-best-errors}
api() |>
  api_post("/process", function(data, res) {
    tryCatch({
      # Process data
      result <- some_processing(data)

      # Success response
      res$status <- 200
      list(success = TRUE, result = result)
    }, error = function(e) {
      # Error response
      res$status <- 400
      list(
        success = FALSE,
        error = "Processing failed",
        details = as.character(e)
      )
    })
  })
```

### 5. Document Your Endpoints

Use comments to describe what each endpoint does:

```{r handler-best-docs}
#* Calculate the sum of two numbers
#*
#* This endpoint performs addition on two numeric values.
#*
#* @param a:numeric* The first number
#* @param b:numeric* The second number
#* @return A JSON object with the sum
#* @get /add
function(a, b) {
  list(a = a, b = b, sum = a + b)
}
```

# Output Formats and Serializers

## Available Serializers

plumber2 supports many output formats out of the box:

```{r list-serializers}
library(plumber2)
show_registered_serializers()
```

| Serializer | MIME Type | Use Case |
|------------|-----------|----------|
| `json` | application/json | Default; structured data |
| `csv` | text/csv | Tabular data, data frames |
| `text` | text/plain | Plain text responses |
| `html` | text/html | HTML content |
| `xml` | text/xml | XML documents |
| `rds` | application/rds | R-specific binary format |
| `png` | image/png | Plots and graphics |
| `pdf` | application/pdf | PDF documents |
| `yaml` | text/yaml | YAML format |
| `feather` | application/vnd.apache.arrow.file | Arrow format |

## JSON Output (Default)

```{r json-output}
api() |>
  api_get("/data", function() {
    list(
      status = "success",
      data = data.frame(
        id = 1:3,
        value = c(10, 20, 30)
      )
    )
  })
# Returns JSON by default
```

## CSV Output

```{r csv-output}
api() |>
  api_get("/export", function() {
    data.frame(
      id = 1:3,
      name = c("Alice", "Bob", "Charlie"),
      score = c(95, 87, 92)
    )
  }, serializers = get_serializers("csv"))
# Returns CSV format
```

## Plain Text Output

```{r text-output}
api() |>
  api_get("/status", function() {
    "Server is running normally"
  }, serializers = get_serializers("text"))
# Returns plain text
```

## Image Output

```{r image-output}
api() |>
  api_get("/plot", function() {
    # Create plot
    plot(1:10, 1:10, main = "Simple Plot")
  }, serializers = get_serializers("png"))
# Returns PNG image
```

## Multiple Serializers

Allow clients to choose format via Accept header:

```{r multiple-serializers}
api() |>
  api_get("/flexible", function() {
    data.frame(x = 1:5, y = 1:5)
  }, serializers = get_serializers(c("json", "csv", "xml")))

# Client specifies: Accept: text/csv
```

## Custom Serializers

Register custom serialization logic:

```{r custom-serializer}
# Define custom serializer
my_serializer <- function(x) {
  paste("CUSTOM:", as.character(x), collapse = ", ")
}

# Register it
register_serializer(
  name = "custom",
  mime_type = "text/custom",
  serializer = my_serializer
)

# Use in endpoint
api() |>
  api_get("/custom", function() {
    c("a", "b", "c")
  }, serializers = get_serializers("custom"))
```

# Hosting the API

## Local Development

```{r local-run}
library(plumber2)

api() |>
  api_get("/hello", function() {
    list(message = "Hello")
  }) |>
  api_run(
    host = "127.0.0.1", # localhost only
    port = 8080,
    showcase = TRUE # Opens browser with API docs
  )
```

**Access:** http://127.0.0.1:8080/hello

## Network Access

Allow connections from other machines:

```{r network-run}
api() |>
  api_get("/hello", function() {
    list(message = "Hello")
  }) |>
  api_run(
    host = "0.0.0.0", # Accept from any IP
    port = 8080
  )
```

**Access:** http://<your-ip>:8080/hello

## RStudio Server

On RStudio Server, use the proxied URL:

```{r rstudio-server}
# Run on specific port
api() |> api_run(port = 8080)

# Access at:
# https://<rstudio-server>/p/8080/
```

## AWS Deployment Options

### 1. EC2 Instance

```bash
# Install R and plumber2
sudo R -e "install.packages('plumber2')"

# Run in background
nohup R -e "library(plumber2); parse_plumber_file('api.R') |> api_run(host='0.0.0.0', port=8080)" &
```

### 2. AWS Lambda

Use `lambdr` package for serverless deployment:

```{r lambda}
# Install lambdr
install.packages("lambdr")

# api.R
library(plumber2)
library(lambdr)

api <- api() |>
  api_get("/hello", function(name = "World") {
    list(greeting = paste("Hello,", name))
  })

lambdr::start_lambda(api)
```

### 3. Docker Container

```dockerfile
# Dockerfile
FROM rocker/r-ver:4.5.1

RUN R -e "install.packages('plumber2')"

COPY api.R /app/api.R
WORKDIR /app

EXPOSE 8080

CMD ["R", "-e", "plumber2::parse_plumber_file('api.R') |> plumber2::api_run(host='0.0.0.0', port=8080)"]
```

```bash
# Build and run
docker build -t my-plumber-api .
docker run -p 8080:8080 my-plumber-api
```

### 4. Elastic Beanstalk

Deploy Docker container to Elastic Beanstalk for managed hosting.

## Production Considerations

```{r production}
api() |>
  api_get("/data", function() {
    # Your logic
  }) |>
  api_run(
    host = "0.0.0.0",
    port = 8080,
    showcase = FALSE, # Disable in production
    silent = TRUE # Reduce logging
  )
```

Use reverse proxy (Nginx) for SSL, load balancing:

```nginx
# nginx.conf
server {
  listen 80;
  server_name api.example.com;

  location / {
    proxy_pass http://localhost:8080;
    proxy_set_header Host $host;
    proxy_set_header X-Real-IP $remote_addr;
  }
}
```

# Configuration with _server.yml

The `_server.yml` file allows centralized configuration of your API.

## Basic Structure

```yaml
# _server.yml
host: 0.0.0.0
port: 8080

# API options
docType: rapidoc        # API documentation style
docPath: __docs__       # Docs endpoint path

# Request handling
maxRequestSize: 10000000  # 10MB limit
compressionLimit: 1000    # Compress responses > 1KB

# Security
sharedSecret: "your-secret-key"

# Logging
logger:
  type: console
  level: info
```

## Loading Configuration

```{r load-config}
library(plumber2)

# Load from _server.yml
api() |>
  api_get("/hello", function() {
    list(message = "Hello")
  }) |>
  api_run() # Automatically reads _server.yml
```

## Environment-Specific Configuration

```yaml
# _server.yml
default:
  host: 127.0.0.1
  port: 8080

production:
  host: 0.0.0.0
  port: 80
  showcase: false

development:
  host: 127.0.0.1
  port: 8080
  showcase: true
```

```{r env-config}
# Set environment
Sys.setenv(R_CONFIG_ACTIVE = "production")

# Loads production settings
api() |> api_run()
```

## Advanced Options

```yaml
# _server.yml
host: 0.0.0.0
port: 8080

# CORS settings
cors:
  enabled: true
  origin: "*"
  methods: ["GET", "POST", "PUT", "DELETE"]

# Rate limiting
rateLimit:
  enabled: true
  requests: 100
  window: 60  # seconds

# Caching
cache:
  enabled: true
  ttl: 300  # seconds

# Async execution
async:
  enabled: true
  evaluator: mirai
```

# Complete Working Example

Here's a comprehensive example demonstrating different input types and output formats:

```{r complete-example}
library(plumber2)

# Create API
demo_api <- api()

# Add multiple endpoints with different methods and outputs
demo_api |>
  # 1. Simple GET with query parameter (JSON response)
  api_get("/hello", function(name = "World") {
    list(
      greeting = paste("Hello,", name),
      timestamp = Sys.time()
    )
  }) |>
  # 2. GET with path parameter
  api_get("/users/{id}", function(id) {
    list(
      user_id = as.integer(id),
      username = paste("user", id, sep = "_"),
      status = "active"
    )
  }) |>
  # 3. POST with body parameters
  api_post("/calculate", function(x, y, operation = "add") {
    x_num <- as.numeric(x)
    y_num <- as.numeric(y)

    result <- switch(operation,
      "add" = x_num + y_num,
      "subtract" = x_num - y_num,
      "multiply" = x_num * y_num,
      "divide" = x_num / y_num,
      NA
    )

    list(
      operation = operation,
      x = x_num,
      y = y_num,
      result = result
    )
  }) |>
  # 4. Plain text response
  api_get("/status", function() {
    paste(
      "Server Status: Running",
      "Time:", Sys.time(),
      "Memory:", format(object.size(ls()), units = "MB"),
      sep = "\n"
    )
  }, serializers = get_serializers("text")) |>
  # 5. CSV export
  api_get("/data/csv", function(rows = 10) {
    data.frame(
      id = 1:as.integer(rows),
      name = paste("Name", 1:as.integer(rows)),
      value = rnorm(as.integer(rows), mean = 100, sd = 15),
      category = sample(c("A", "B", "C"), as.integer(rows), replace = TRUE)
    )
  }, serializers = get_serializers("csv")) |>
  # 6. JSON data export
  api_get("/data/json", function(rows = 10) {
    data.frame(
      id = 1:as.integer(rows),
      name = paste("Name", 1:as.integer(rows)),
      value = rnorm(as.integer(rows), mean = 100, sd = 15),
      category = sample(c("A", "B", "C"), as.integer(rows), replace = TRUE)
    )
  }) |>
  # 7. Image response (plot)
  api_get("/plot", function(n = 10) {
    n <- as.integer(n)
    x <- 1:n
    y <- rnorm(n, mean = x, sd = 5)

    plot(x, y,
      main = "Random Data Plot",
      xlab = "X Values",
      ylab = "Y Values",
      pch = 19,
      col = "steelblue",
      cex = 1.5
    )
    lines(x, x, col = "red", lwd = 2)
  }, serializers = get_serializers("png")) |>
  # 8. Multiple output formats (client chooses via Accept header)
  api_get("/flexible", function() {
    data.frame(
      metric = c("CPU", "Memory", "Disk"),
      usage = c(45.2, 67.8, 23.5),
      unit = c("%", "%", "%")
    )
  }, serializers = get_serializers(c("json", "csv", "xml"))) |>
  # 9. Access request headers
  api_get("/headers", function(req) {
    list(
      user_agent = req$HTTP_USER_AGENT,
      accept = req$HTTP_ACCEPT,
      host = req$HTTP_HOST,
      all_headers = names(req)
    )
  }) |>
  # 10. Error handling
  api_get("/divide", function(a, b) {
    a_num <- as.numeric(a)
    b_num <- as.numeric(b)

    if (b_num == 0) {
      # Return error response
      list(
        error = TRUE,
        message = "Division by zero is not allowed"
      )
    } else {
      list(
        error = FALSE,
        result = a_num / b_num
      )
    }
  })

# Print endpoint summary
cat("API created with the following endpoints:\n")
cat("\nGET endpoints:\n")
cat("  /hello?name=<name>           - Greeting with name\n")
cat("  /users/{id}                  - Get user by ID\n")
cat("  /status                      - Server status (plain text)\n")
cat("  /data/csv?rows=<n>           - Export data as CSV\n")
cat("  /data/json?rows=<n>          - Export data as JSON\n")
cat("  /plot?n=<number>             - Generate plot (PNG)\n")
cat("  /flexible                    - Multiple formats (JSON/CSV/XML)\n")
cat("  /headers                     - Show request headers\n")
cat("  /divide?a=<num>&b=<num>      - Division with error handling\n")
cat("\nPOST endpoints:\n")
cat("  /calculate                   - Calculate operations\n")
cat("    Body: {\"x\": 10, \"y\": 5, \"operation\": \"add\"}\n")
cat("\n")

# Run the server (uncomment to start)
# demo_api |> api_run(port = 8080, showcase = TRUE)
```

## Testing the API

Once running, test with curl or R:

```{r test-api, eval=FALSE}
library(httr2)

# Test 1: Simple GET
request("http://localhost:8080/hello") |>
  req_url_query(name = "Alice") |>
  req_perform() |>
  resp_body_json()

# Test 2: POST request
request("http://localhost:8080/calculate") |>
  req_body_json(list(x = 10, y = 5, operation = "multiply")) |>
  req_perform() |>
  resp_body_json()

# Test 3: Get CSV
response <- request("http://localhost:8080/data/csv") |>
  req_url_query(rows = 5) |>
  req_perform()

df <- response |>
  resp_body_string() |>
  read.csv(text = _)
print(df)

# Test 4: Download plot
request("http://localhost:8080/plot") |>
  req_url_query(n = 20) |>
  req_perform() |>
  resp_body_raw() |>
  writeBin("plot.png")
```

Or with curl:

```bash
# Simple GET
curl "http://localhost:8080/hello?name=Alice"

# POST with JSON
curl -X POST http://localhost:8080/calculate \
  -H "Content-Type: application/json" \
  -d '{"x": 10, "y": 5, "operation": "add"}'

# Get CSV
curl "http://localhost:8080/data/csv?rows=5"

# Download plot
curl "http://localhost:8080/plot?n=20" --output plot.png

# Flexible format (CSV via Accept header)
curl "http://localhost:8080/flexible" \
  -H "Accept: text/csv"
```

# Best Practices

## 1. Error Handling

```{r error-handling}
api() |>
  api_get("/safe", function(x) {
    tryCatch(
      {
        result <- as.numeric(x) * 2
        list(success = TRUE, result = result)
      },
      error = function(e) {
        list(
          success = FALSE,
          error = as.character(e)
        )
      }
    )
  })
```

## 2. Authentication

```{r auth}
api() |>
  api_auth(function(req) {
    token <- req$HTTP_AUTHORIZATION
    if (is.null(token) || token != "Bearer secret123") {
      return(FALSE)
    }
    TRUE
  }) |>
  api_get("/protected", function() {
    list(message = "Access granted")
  })
```

## 3. Request Validation

```{r validation}
api() |>
  api_post("/submit", function(email, age) {
    # Validate inputs
    if (!grepl("@", email)) {
      return(list(
        error = TRUE,
        message = "Invalid email format"
      ))
    }

    age_num <- as.numeric(age)
    if (is.na(age_num) || age_num < 0 || age_num > 150) {
      return(list(
        error = TRUE,
        message = "Invalid age"
      ))
    }

    list(
      error = FALSE,
      message = "Submission accepted",
      data = list(email = email, age = age_num)
    )
  })
```

## 4. Logging

```{r logging}
api() |>
  api_logger(function(req, res) {
    cat(sprintf(
      "[%s] %s %s - %s\n",
      Sys.time(),
      req$REQUEST_METHOD,
      req$PATH_INFO,
      res$status
    ))
  }) |>
  api_get("/logged", function() {
    list(message = "This request is logged")
  })
```

## 5. CORS for Web Apps

```{r cors}
api() |>
  api_security_cors(
    origins = c("http://localhost:3000", "https://myapp.com"),
    methods = c("GET", "POST", "PUT", "DELETE"),
    headers = c("Content-Type", "Authorization")
  ) |>
  api_get("/data", function() {
    list(data = "Available to web apps")
  })
```

# Summary

| Topic | Key Points |
|-------|------------|
| **Purpose** | Create REST APIs to expose R functions as web services |
| **vs Shiny** | APIs for programmatic access; Shiny for interactive UIs |
| **Creation** | Functional (`api_get()`) or annotation-based (`#* @get`) |
| **Methods** | GET, POST, PUT, DELETE, PATCH |
| **Inputs** | Query params, path params, body, headers |
| **Outputs** | JSON, CSV, text, XML, images, and more |
| **Hosting** | Local, RStudio Server, EC2, Lambda, Docker |
| **Config** | `_server.yml` for centralized settings |
| **Production** | Add auth, validation, logging, CORS, error handling |

# Resources

- **Official Documentation:** https://plumber2.posit.co/
- **Vignettes:** https://plumber2.posit.co/articles/
- **GitHub:** https://github.com/rstudio/plumber2
- **Original plumber:** For comparison and migration guides
