---
title: "Reticulate: calling python in R"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
  github_document:
    toc: true
---

```{r setup, include=F}
library(knitr)
library(reticulate)
# set the python to use
use_python("d:/Tools/Python/Python38/python.exe")
opts_chunk$set(echo=T, comment=">")
```

## Introduction

[reticulate](https://rstudio.github.io/reticulate/) 
is a package enabling embedding python session in R,
which provides convenience for using some python packages which
don't have R equivalents.

Normally, the conversion between R and python objects takes place
automatically. So one can access the variables from the other language
easily.

## Setup

To call python in R, one need some settings:

1. install 'reticulate' R package.

  ```r
  install.packages("reticulate")
  ```

2. choose the python to use. One can set the right python using the functions
	such as `use_python()`, `use_python_version()`, `use_condaenv()`, `use_miniconda()`.
	If this is not set beforehand, it will prompt you to install miniconda, even though
	you have installed pythons somewhere. This setup needs be done only once for one account
	in a machine. See [here](https://rstudio.github.io/reticulate/articles/versions.html)
	for more details.

## 4 ways to call python

1. Python in R Markdown — A new Python language engine for R Markdown that supports bi-directional communication between R and Python (R chunks can access Python objects and vice-versa).

2. Importing Python modules — The import() function enables you to import any Python module and call it’s functions directly from R.

3. Sourcing Python scripts — The source_python() function enables you to source a Python script the same way you would source() an R script (Python functions and objects defined within the script become directly available to the R session).

4. Python REPL — The repl_python() function creates an interactive Python console within R. Objects you create within Python are available to your R session (and vice-versa).

See [this page](https://rstudio.github.io/reticulate/) for details.


## Python in Rmarkdown

In this mode, reticulate includes a Python engine for Rmarkdown that enables
convenient sharing R and Python objects. For example, the NumPy arrays and pandas
DataFrame are internally converted into vector and data.frame when accessing them
in R chunks. **Note**, one need have all the packages such as pandas and NumPy
installed to get correct object conversion, otherwise, the objects may be converted
to something else (e.g., R data.frame to python dictionary).

Python chunks all execute within a single Python session so have access to all
objects created in previous chunks. 

### Run python chunks

Below is running python code in R

```{python}
s="This is a string printed from Python in Rmarkdown"
print(s)
s+=". Welcome (edited)!"
print(s)
```

### Access R variables in Python

First, let's summary the data in R.

```{r}
cat("The R object `mtcars` is ", class(mtcars),"\n")
summary(mtcars)
```

Now, let's make a boxplot with the data.

```{python}
import matplotlib.pyplot as plt
import pandas as pd
dat=r.mtcars
print(f"I got an R object and converted to type {type(dat)}")
dat.head()
bplot=dat.boxplot(column="mpg", by="gear", color="blue", grid=False)
tmp=bplot.plot()
plt.title("Boxplot in Python")
plt.show()
```


### Access Python variables in R

Let's create a pandas.DataFrame first.

```{python}
import pandas as pd
import numpy as np
# create a random number list
df=pd.DataFrame({'x':np.random.rand(30), 'g':['a','b','c']*10})
df.head()
```

Now, let's read this data frame into R

```{r}
dat<-py$df
cat("I got an python object 'df' in R with type: ", class(dat), "\n")
summary(dat)
library(ggplot2)
plt<-ggplot(dat, aes(x=x)) + geom_density()
plt + facet_wrap(~g)
```

### Exchanging variables between R and Python

Now let's create a list of variables of different types in R and see what their
types are in Python.

```{r}
rVector1<-13
rVector2<-1:10
rList1<-list(1:3)
rList2<-list(a=1:2,b="hello")
rMatrix<-matrix(1:3)
rDF<-data.frame()
rFun<-function() {print("R function")}
rLogical<-TRUE
rNull<-NULL
rRaw<-raw(3)
```

Here are the types that Python received:

```{python}
print(f"""
R vector of length 1: {type(r.rVector1)}
R vector of length >1: {type(r.rVector2)}
R list: {type(r.rList1)}
R named list: {type(r.rList2)}
R matrix: {type(r.rMatrix)}
R data.frame: {type(r.rDF)}
R function: {type(r.rFun)}
R logical: {type(r.rLogical)}
R NULL: {type(r.rNull)}
R raw: {type(r.rRaw)}
""")
```

## Import Python modules

One can use the function [reticulate::import()] to import python modules and use
the module's functions and properties using the symbol '$'. See the following examples:

```{r}
np=import("numpy")
print(class(np))
a=np$array(2:4)
print(a)
print(sum(a))
```

Note that if one specify the parameter 'convert=F' when importing a python object,
for example, a function, then the object and its derivatives are in python format,
which can't use R's functions. To convert an object manually, use the function
[reticulate::py_to_r()].


```{r}
np=import("numpy", convert = F)
print(class(np))
a=np$array(2:4)
print(a)
tryCatch(print(sum(a)), error=function(e) { print("Invalid argument for R function sum()") })
a=py_to_r(a)
print(paste("After manual conversion:", sum(a)))
```


### Summary of Python and R object conversion

The corresponding objects converted from R to Python and 
vice versa.

R object | python object
--- | ---
length-one vector | scalar 
vector longer than 1 | List 
List | Tuple  
named list | Dict 
Matrix/Array | NumPy ndarray 
data.frame | Pandas DataFrame 
Raw | bytearray 
Function | Python function  
NULL | None 
TRUE/FALSE | True/False
Factor | Categorical variable
POSIXt | NumPy array with dtype datetime64[ns]

## Source python code

One can source a python script as sourcing an R script, and use the functions and
objects generated in the python script directly. For example:

```python
# python script: call_python.py
global_x=100
def py_add(a,b):
  return a+b
```

```{r}
source_python("static/call_python.py")
cat("I see python variable global_x as: ", global_x,"\n")
# now let's call the function
res<-py_add(1,3)
cat("Calling the python function py_add(1,3), I got: ", res, "\n")
```


## Use Python REPL

Python REPL is an embeded python session into an R session. Objects created within
Python REPL can be accessed in R using the object `py`.

This mode is only available when working interactively, triggered by typing
the function `repl_python()`, and one can exit to R session
by typing `exit`.

## Important functions

### Functions to import python objects into R

Function | Use
--- | ---
py_to_r() | convert python object to R equivalent
py_func() | define a python function using the R syntax, like `my_py_fun<-py_func(function(a, b = 1.5) {})`.
r_to_py() | convert R object to python equivalent
souce_python() | read all python functions in the sourced file, and these functions can be called directly, like R functions
import() | import python module, and access the functions/variables using `$` attribute symbol
import_builtins() | import python's bultin functions
import_main() | import main modules, useful for generating some objects in python and accessing them from R using `py$x` format.
repl_python() | create an interactive python console within R, using `py$x` and `r.x` to access the variable `x` in python and R space, respectively
py_help() | print help information on python object
py_get_attr() | get attribute of python object, no python-to-R conversion by this function, in contrast to `py$x` operation.
py_call() | call a python callable with arguments
py_len() | get the length of a python object.
py_save_object() | save a python object to file
py_load_object() | load saved python object from file.
py_config() | get python configuration
tuple() | create a Python tuple
dict() | create a python dictionary
np_array() | create a numpy array
py_iterator() | convert R function to python generator
py_func() | convert an R function to python function, keeping the same signatures such as default values.

### Functions that check Python objects without conversion

These functions can check a Python object's properties without conversion to R
object.

Function | Use
--- | :---
py_has_attr() 	Check if an object has a specified attribute.
py_get_attr() 	Get an attribute of a Python object.
py_set_attr() 	Set an attribute of a Python object.
py_list_attributes() 	List all attributes of a Python object.
py_len() 	Length of Python object.
py_call() 	Call a Python callable object with the specified arguments.

### Functions to check Python setting

Function | Use
--- | :---
py_available() 	Check whether a Python interface is available on this system.
py_numpy_available() 	Check whether the R interface to NumPy is available (requires NumPy >= 1.6)
py_module_available() 	Check whether a Python module is available on this system.
py_config() 	Get information on the location and version of Python in use.

### Other useful functions

Function | Use
--- | :---
py_capture_output() |	Capture Python output for the specified expression and return it as an R character vector.
py_suppress_warnings() |	Execute the specified expression, suppressing the display Python warnings.
py_set_seed() |	Set Python and NumPy random seeds.
py_unicode() |	Convert a string to a Python unicode object.
py_str(), py_repr() |	Get the string representation of Python object.
py_id() |	Get a unique identifier for a Python object
py_is_null_xptr() |	Check whether a Python object is a null externalptr.
py_validate_xptr() |	Check whether a Python object is a null externalptr and throw an error if it is.


## Release memory used by Python

To release memory by python object, one need do the following:

1. delete the object by `py_run_string("del pyobj")`

2. then collect the garbage with `py_gc <- import("gc"); py_gc$collect()`


## References

1. Calling python in R: https://rstudio.github.io/reticulate/articles/calling_python.html

2. reticulate homepage: https://rstudio.github.io/reticulate/
