---
title: "How to use shinychat with ellmer"
author: "Zhenguo Zhang"
date: "`r Sys.Date()`"
# use github format
output:
  github_document:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings=FALSE, message=FALSE)
workDir<-tempdir()
```

## Introduction

In this tutorial, we will learn how to use the `shinychat` package in
R to create interactive chat applications. The `shinychat` package
allows you to build chat interfaces that can be integrated with
various AI models, including those provided by OpenAI. We will go
through the steps of setting up a simple chat application using
`shinychat` and how to connect it with an AI model for generating
responses.

## Setting Up the Environment

To get started, you need to install the `shinychat` package. You can
do this using the following command:

```{r install, eval=FALSE}
install.packages("shinychat")
```

Once you have the package installed, you can load it into your R
session:
```{r load}
library(shinychat)
```

## Creating a Simple Chat application

To create a simple chat application, you can use the function
`chat_mod_ui()` to define the user interface and `chat_mod_server()`
to define the server logic. Here’s a basic example of how to set up a
chat application:

```{r chat-app, eval=FALSE}
library(shiny)
library(shinychat)
library(ellmer)

ui <- fluidPage(
  chat_mod_ui("chat1")
)

server <- function(input, output, session) {
    # create a chat client using ellmer
    chat_client <- ellmer::chat_google_gemini(model =
                                              "gemini-2.5-flash")
  chat_mod_server("chat1", client = chat_client)
}

shinyApp(ui, server)
```

In this example, we create a simple Shiny application with a chat
interface. We use the `chat_mod_ui()` function to create the user
interface for the chat and the `chat_mod_server()` function to handle
the server logic. We also create a chat client using the `ellmer`
package, which allows us to connect to the Google Gemini model for
generating responses.

## Tips and pitfalls

### Critical: Module ID Management

If you call `chat_mod_server()` multiple times, make sure to use
different module IDs for each instance. For example, you can use
"chat1", "chat2", etc. This is crucial to avoid conflicts between
different chat instances, and to avoid the error of ID conflicts in
shiny modules.

#### ❌ Don't: Reuse the same ID

```r
# WRONG: This causes UI conflicts when switching models
chat_client <- chat_mod_server("my_chat", client = new_client)
# Later, trying to create another instance with the same ID...
chat_client <- chat_mod_server("my_chat", client = different_client)
```

**Problem**: Shiny's module system requires unique IDs for each instance. Reusing IDs causes:
- UI elements from old instances persist and interfere
- Event handlers from previous instances continue firing
- Chat history and state get mixed between instances

#### ✅ Do: Use dynamic, unique IDs

```r
# CORRECT: Generate a unique ID for each instance
new_chat_id <- paste0("chat_", paste(sample(c(0:9, letters, LETTERS), 15, replace = TRUE), collapse = ""))
chat_client <- chat_mod_server(new_chat_id, client = new_client)
```

### Dynamic UI Rendering Pattern

#### ❌ Don't: Hardcode module IDs in UI

```r
# WRONG: Static ID in UI
chat_mod_ui("my_chat")
```

#### ✅ Do: Use renderUI() for dynamic module creation

```r
# In UI:
uiOutput("chat_ui")

# In server:
chat_id <- reactiveVal(NULL)
# when the chat_id changes, the UI will automatically update
output$chat_ui <- renderUI({
  req(chat_id)
  chat_mod_ui(chat_id)
})
```

**Why this works**: When `chat_id` changes, `renderUI()` automatically:
1. Removes the old module UI
2. Creates new module UI with the new ID
3. Cleans up event handlers from the old module

