---
title: "Learning tidymodels: A Modern Framework for Modeling in R"
author: "Learning Notes"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction to tidymodels

`tidymodels` is a collection of packages for modeling and machine learning using tidyverse principles. It provides a consistent, modular framework for:

- Data preprocessing and feature engineering
- Model specification and training
- Hyperparameter tuning
- Model evaluation and comparison
- Workflow management

Note that Tidymodels itself doesn’t implement any statistical or machine learning models. It’s a framework to organize all the tasks around modeling.


## Core Packages

```{r load-packages}
library(tidymodels) # Loads the core tidymodels packages
library(tidyverse) # For data manipulation

# Core packages included in tidymodels:
# - rsample: data splitting and resampling
# - recipes: data preprocessing and feature engineering
# - parsnip: unified model interface
# - workflows: combine preprocessing and modeling
# - tune: hyperparameter tuning
# - yardstick: model performance metrics
# - broom: tidy model outputs
# - dials: hyperparameter space definitions
```

---

# The tidymodels Workflow

A typical tidymodels workflow follows these steps:

1. **Split data** (rsample)
2. **Create recipe** for preprocessing (recipes)
3. **Specify model** (parsnip)
4. **Create workflow** (workflows)
5. **Fit model** or **tune hyperparameters** (tune)
6. **Evaluate performance** (yardstick)

---

# 1. Data Splitting with rsample

## Basic Train/Test Split

```{r data-split}
# Example dataset
data(ames, package = "modeldata")

# Set seed for reproducibility
set.seed(123)

# Split data: 75% training, 25% testing
ames_split <- initial_split(ames, prop = 0.75, strata = Sale_Price)

ames_train <- training(ames_split)
ames_test <- testing(ames_split)

cat("Training set size:", nrow(ames_train), "\n")
cat("Testing set size:", nrow(ames_test), "\n")
```

## Cross-Validation

```{r cross-validation}
# 10-fold cross-validation
ames_folds <- vfold_cv(ames_train, v = 10, strata = Sale_Price)

# Bootstrap resampling
ames_boots <- bootstraps(ames_train, times = 25, strata = Sale_Price)

# Time-series cross-validation
# For time series data
# time_slices <- rolling_origin(data, initial = 365, assess = 30, cumulative = TRUE)
```

## Validation Set

```{r validation-set}
# Create train/validation/test split
set.seed(456)

# First split: separate test set
ames_split2 <- initial_split(ames, prop = 0.8, strata = Sale_Price)
ames_other <- training(ames_split2)
ames_test2 <- testing(ames_split2)

# Second split: separate training and validation
ames_val_split <- initial_split(ames_other, prop = 0.75, strata = Sale_Price)
ames_train2 <- training(ames_val_split)
ames_val <- testing(ames_val_split)
```

## Key Methods for rsample Objects

Understanding the methods available for different rsample objects helps you work efficiently with data splits and resamples.

### For `initial_split` objects:

| Method | Description | Returns |
|--------|-------------|---------|
| `training(split)` | Extract training data | Data frame/tibble |
| `testing(split)` | Extract test data | Data frame/tibble |
| `complement(split)` | Get indices not in training | Integer vector |

```{r rsample-split-methods, eval=FALSE}
split <- initial_split(data, prop = 0.75)
train <- training(split)  # Get training set
test <- testing(split)    # Get test set
```

### For resampling objects (`vfold_cv`, `bootstraps`, etc.):

| Method | Description | Returns |
|--------|-------------|---------|
| `analysis(resample)` | Extract analysis/training fold | Data frame |
| `assessment(resample)` | Extract assessment/validation fold | Data frame |
| `rsample2caret(folds)` | Convert to caret format | List of indices |
| `pretty.rsplit()` | Print split information | Text summary |

```{r rsample-cv-methods, eval=FALSE}
folds <- vfold_cv(data, v = 10)

# Access individual fold
fold1 <- folds$splits[[1]]
fold1_train <- analysis(fold1)      # Training portion of fold 1
fold1_validate <- assessment(fold1) # Validation portion of fold 1

# Convert for caret compatibility
caret_indices <- rsample2caret(folds)
```

### Common resampling functions:

| Function | Purpose | Key Parameters |
|----------|---------|----------------|
| `vfold_cv()` | V-fold cross-validation | `v` (# folds), `repeats`, `strata` |
| `bootstraps()` | Bootstrap resampling | `times` (# bootstrap samples), `strata` |
| `mc_cv()` | Monte Carlo CV | `prop`, `times` |
| `rolling_origin()` | Time series CV | `initial`, `assess`, `cumulative` |
| `group_vfold_cv()` | Grouped CV | `group`, `v` |
| `loo_cv()` | Leave-one-out CV | None (one-per-observation) |

## Cross-Package Compatibility

If you need to compare tidymodels with caret using identical resamples:

```{r rsample-caret-compat, eval=FALSE}
# Create tidymodels folds
tidy_folds <- vfold_cv(ames_train, v = 10, strata = Sale_Price)

# Convert to caret format for fair comparison
caret_folds <- rsample2caret(tidy_folds)

# Now both packages use identical data splits
tidy_results <- workflow %>% fit_resamples(resamples = tidy_folds)
caret_results <- train(..., trControl = trainControl(index = caret_folds))

# Convert back if needed
back_to_tidy <- caret2rsample(caret_folds)
```

**Why this matters**: When benchmarking frameworks, using identical resamples ensures performance differences are due to the framework, not random variation in splits.

---

# 2. Feature Engineering with recipes

Recipes define preprocessing steps that are applied consistently to training and test data.

## Basic Recipe

```{r basic-recipe}
# Create a simple recipe
ames_recipe <- recipe(Sale_Price ~ ., data = ames_train) %>%
  # Remove zero variance predictors
  step_zv(all_predictors()) %>%
  # Normalize numeric predictors
  step_normalize(all_numeric_predictors()) %>%
  # Create dummy variables for factors
  step_dummy(all_nominal_predictors())

# View recipe
ames_recipe
```

## Common Recipe Steps

```{r recipe-steps}
advanced_recipe <- recipe(Sale_Price ~ ., data = ames_train) %>%
  # Handle missing data
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  # Feature engineering
  step_log(Sale_Price, base = 10) %>% # Log transform outcome
  step_date(all_date_predictors(), features = c("dow", "month", "year")) %>%
  # Handle categorical variables
  step_other(all_nominal_predictors(), threshold = 0.01) %>% # Pool rare levels
  step_dummy(all_nominal_predictors(), one_hot = FALSE) %>%
  # Handle correlations and variance
  step_zv(all_predictors()) %>%
  step_corr(all_numeric_predictors(), threshold = 0.9) %>%
  # Normalization (choose one)
  step_normalize(all_numeric_predictors()) %>% # Center and scale
  # step_range(all_numeric_predictors(), min = 0, max = 1) %>%  # Scale to range

  # Interaction terms
  step_interact(~ Gr_Liv_Area:starts_with("Bldg_Type"))

# Prep and bake to see the result
advanced_recipe %>%
  prep() %>%
  bake(new_data = NULL) %>%
  glimpse()
```

## Important Recipe Concepts

### Recipes as Blueprints

**Key insight**: A recipe is a *blueprint* that defines your preprocessing steps **before their actual execution**. This ensures:
- Preprocessing is consistent between training and test data
- Statistics (means, SDs, etc.) are computed only on training data
- No data leakage occurs

```{r recipe-concept, eval=FALSE}
# Recipe creation (just defines the steps)
my_recipe <- recipe(y ~ ., data = train) %>%
  step_normalize(all_numeric_predictors()) # Doesn't normalize yet!

# Prepare recipe (calculates statistics from training data)
prepped_recipe <- prep(my_recipe, training = train)

# Apply recipe (uses training statistics)
train_processed <- bake(prepped_recipe, new_data = train)
test_processed <- bake(prepped_recipe, new_data = test) # Uses train stats!
```

### Variable Roles

Tidymodels allows assigning roles to variables - useful for keeping IDs without using them in modeling:

```{r variable-roles, eval=FALSE}
recipe(outcome ~ ., data = train) %>%
  update_role(employee_id, new_role = "id") %>% # Keep but don't model
  step_normalize(all_numeric_predictors())

# employee_id is retained but excluded from preprocessing and modeling
```

### Recipe Role Selectors

Recipes provides powerful selector functions to target specific groups of variables based on their role and type. These are essential for writing flexible, maintainable preprocessing code.

| Selector Function | Description | Example Use Case |
|------------------|-------------|------------------|
| `all_predictors()` | All predictor variables (not outcomes) | Apply transformation to all features |
| `all_outcomes()` | All outcome/target variables | Transform the response variable |
| `all_numeric()` | All numeric variables (including outcomes) | Select all numeric columns |
| `all_nominal()` | All categorical variables (factors/characters) | Select all categorical columns |
| `all_numeric_predictors()` | Numeric predictor variables only | Normalize or scale features |
| `all_nominal_predictors()` | Categorical predictor variables only | Create dummy variables |
| `has_role()` | Variables with a specific role | Target custom roles like "ID" |
| `has_type()` | Variables of a specific type | Target specific data types |

**Type-based selectors:**

| Selector Function | Selects |
|------------------|---------|
| `all_integer()` | Integer columns |
| `all_double()` | Double/numeric columns |
| `all_logical()` | Logical/boolean columns |
| `all_string()` | Character/string columns |
| `all_factor()` | Factor columns |
| `all_ordered()` | Ordered factor columns |
| `all_date()` | Date columns |
| `all_datetime()` | Datetime columns |

**Combining selectors:**

```{r selector-examples, eval=FALSE}
recipe(Sale_Price ~ ., data = train) %>%
  # Normalize all numeric predictors
  step_normalize(all_numeric_predictors()) %>%

  # Create dummies for all categorical predictors
  step_dummy(all_nominal_predictors()) %>%

  # Log transform outcome and specific predictors
  step_log(all_outcomes(), Lot_Area, Gr_Liv_Area) %>%

  # Remove zero variance from all predictors
  step_zv(all_predictors()) %>%

  # Use tidyselect helpers
  step_rm(starts_with("temp_")) %>%
  step_interact(~ Gr_Liv_Area:ends_with("_Area"))
```

**Using `has_role()` for custom roles:**

```{r custom-roles, eval=FALSE}
recipe(Sale_Price ~ ., data = train) %>%
  # Assign custom roles
  update_role(property_id, new_role = "id") %>%
  update_role(sale_date, new_role = "date_variable") %>%

  # Target variables by custom role
  step_rm(has_role("id")) %>%  # Remove ID variables before modeling
  step_date(has_role("date_variable"), features = c("month", "year"))
```

**Selector precedence and specificity:**

- Selectors are evaluated in order within each step
- More specific selectors override general ones when combined
- Use `-` to exclude: `step_normalize(all_numeric_predictors(), -Sale_Price)`

### Handling Class Imbalance in Recipes

Use the `themis` package for oversampling/undersampling:

```{r class-imbalance, eval=FALSE}
library(themis)

balanced_recipe <- recipe(churn ~ ., data = train) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_rose(churn) # ROSE algorithm for synthetic oversampling

# Other options:
# step_upsample() - random oversampling
# step_downsample() - random undersampling
# step_smote() - SMOTE algorithm
# step_adasyn() - ADASYN algorithm
```

## Key Methods for Recipe Objects

Recipe objects have specific methods for preparation, application, and inspection.

| Method | Description | When to Use | Returns |
|--------|-------------|-------------|---------|
| `prep(recipe)` | Calculate statistics from training data | After defining recipe, before applying | Prepped recipe |
| `bake(recipe, new_data)` | Apply recipe to new data | After prep(), for test/new data | Processed data |
| `juice(recipe)` | Extract processed training data | After prep(), shortcut for training data | Processed data |
| `tidy(recipe)` | Extract step information | Inspect what each step does | Tibble of steps |
| `summary(recipe)` | Variable summary | See variable roles and types | Tibble of variables |
| `update_role()` | Change variable roles | Assign custom roles (e.g., "ID") | Updated recipe |
| `step_*()` | Add preprocessing steps | Building the recipe | Updated recipe |

```{r recipe-methods, eval=FALSE}
# Create recipe
my_recipe <- recipe(y ~ ., data = train) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())

# Prepare (fit to training data)
prepped_recipe <- prep(my_recipe, training = train)

# Apply to data
train_processed <- juice(prepped_recipe)              # Training data (cached)
test_processed <- bake(prepped_recipe, new_data = test) # Test data

# Inspect recipe
summary(prepped_recipe)       # See all variables and their roles
tidy(prepped_recipe)          # See all steps
tidy(prepped_recipe, number = 1)  # See specific step details
```

### Important `prep()` arguments:

| Argument | Description | Default |
|----------|-------------|---------|
| `training` | Data to calculate statistics from | Required |
| `fresh` | Recalculate from scratch? | `FALSE` |
| `verbose` | Print progress? | `FALSE` |
| `retain` | Keep processed training data? | `TRUE` |
| `strings_as_factors` | Convert strings to factors? | `TRUE` |

### Important `bake()` arguments:

| Argument | Description | Default |
|----------|-------------|---------|
| `new_data` | Data to process (use `NULL` for training) | Required |
| `composition` | How to return data ("tibble", "matrix", "dgCMatrix") | "tibble" |
| `...` | Specific variables to return | All variables |

```{r bake-examples, eval=FALSE}
# Bake specific columns only
bake(prepped_recipe, new_data = test, Sale_Price, Gr_Liv_Area)

# Return as matrix for some models
bake(prepped_recipe, new_data = test, composition = "matrix")
```

## Recipe Step Categories

```{r recipe-reference, eval=FALSE}
# Data cleaning
step_filter() # Filter rows
step_select() # Select columns
step_naomit() # Remove rows with NAs
step_zv() # Remove zero variance
step_nzv() # Remove near-zero variance

# Missing data
step_impute_mean() # Impute with mean
step_impute_median() # Impute with median
step_impute_mode() # Impute with mode
step_impute_knn() # KNN imputation
step_impute_bag() # Bagged tree imputation

# Transformations
step_log() # Log transformation
step_sqrt() # Square root
step_BoxCox() # Box-Cox transformation
step_YeoJohnson() # Yeo-Johnson transformation

# Categorical variables
step_dummy() # Create dummy variables
step_other() # Pool infrequent levels
step_novel() # Handle new factor levels in test data
step_unknown() # Assign missing to "unknown" level

# Normalization
step_normalize() # Center and scale (z-score)
step_center() # Center only
step_scale() # Scale only
step_range() # Scale to min-max range

# Feature engineering
step_poly() # Polynomial features
step_interact() # Interaction terms
step_date() # Extract date features
step_holiday() # Create holiday indicators
step_pca() # Principal component analysis
step_ica() # Independent component analysis

# Text processing
step_tokenize() # Tokenize text
step_stopwords() # Remove stop words
step_stem() # Stem words
step_tf() # Term frequency
step_tfidf() # TF-IDF
```

---

# 3. Model Specification with parsnip

Parsnip provides a unified interface to many modeling engines.

## Linear Regression

```{r linear-regression}
# Specify linear regression model
lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

# Alternative engines
# lm_spec_glmnet <- linear_reg(penalty = 0.1, mixture = 0.5) %>%
#   set_engine("glmnet")
#
# lm_spec_stan <- linear_reg() %>%
#   set_engine("stan")
```

## Elastic Net Regression (Regularized Regression)

Elastic Net combines L1 (Lasso) and L2 (Ridge) regularization, controlled by the `mixture` parameter.

```{r elastic-net}
# Elastic Net with glmnet engine
# mixture: 0 = Ridge, 1 = Lasso, 0.5 = Elastic Net
# penalty: amount of regularization (lambda)

# Ridge Regression (L2 only)
ridge_spec <- linear_reg(penalty = 0.1, mixture = 0) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

# Lasso Regression (L1 only)
lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

# Elastic Net (mix of L1 and L2)
elastic_spec <- linear_reg(penalty = 0.1, mixture = 0.5) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

# For tuning penalty and mixture
elastic_tune_spec <- linear_reg(
  penalty = tune(),
  mixture = tune()
) %>%
  set_engine("glmnet") %>%
  set_mode("regression")
```

### When to Use Each Regularization Method

```{r regularization-guide, eval=FALSE}
# Ridge (mixture = 0):
# - Good when all predictors are relevant
# - Shrinks coefficients but keeps all variables
# - Better for multicollinearity
# - Use when: p < n and you want to keep all features

# Lasso (mixture = 1):
# - Good for feature selection
# - Shrinks some coefficients to exactly zero
# - Creates sparse models
# - Use when: you have many irrelevant features

# Elastic Net (mixture = 0.1 to 0.9):
# - Balances feature selection and coefficient shrinkage
# - Good when features are correlated
# - More stable than Lasso with correlated predictors
# - Use when: p > n or highly correlated features

# Tuning example with elastic net
elastic_recipe <- recipe(Sale_Price ~ ., data = ames_train) %>%
  step_log(Sale_Price) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())

elastic_workflow <- workflow() %>%
  add_recipe(elastic_recipe) %>%
  add_model(elastic_tune_spec)

# Create tuning grid
elastic_grid <- grid_regular(
  penalty(range = c(-5, 0)), # 10^-5 to 10^0
  mixture(range = c(0, 1)),
  levels = c(penalty = 10, mixture = 5)
)

# Tune
elastic_results <- elastic_workflow %>%
  tune_grid(
    resamples = ames_folds,
    grid = elastic_grid,
    metrics = metric_set(rmse, rsq)
  )

# View best parameters
elastic_results %>%
  show_best(metric = "rmse")

# Visualize
elastic_results %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = penalty, y = mean, color = factor(mixture))) +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  labs(
    title = "Elastic Net Tuning Results",
    x = "Penalty (lambda)",
    y = "RMSE",
    color = "Mixture (alpha)"
  )
```

## Random Forest

```{r random-forest}
# Random forest for regression
rf_spec <- rand_forest(
  trees = 1000,
  min_n = 5
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

# For classification
rf_class_spec <- rand_forest(
  trees = 1000,
  min_n = 5
) %>%
  set_engine("ranger") %>%
  set_mode("classification")
```

## Boosted Trees (XGBoost)

```{r xgboost}
# XGBoost specification
xgb_spec <- boost_tree(
  trees = 1000,
  tree_depth = 6,
  learn_rate = 0.01,
  loss_reduction = 0,
  sample_size = 0.8,
  mtry = 10,
  min_n = 5
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")
```

## Other Common Models

```{r other-models}
# Logistic regression
logistic_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Decision tree
tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("regression")

# Support vector machine
svm_spec <- svm_rbf() %>%
  set_engine("kernlab") %>%
  set_mode("regression")

# K-nearest neighbors
knn_spec <- nearest_neighbor(neighbors = 5) %>%
  set_engine("kknn") %>%
  set_mode("regression")

# Neural network
nnet_spec <- mlp(hidden_units = 10, epochs = 100) %>%
  set_engine("nnet") %>%
  set_mode("regression")
```

## Key Methods for Parsnip Model Objects

Parsnip model specifications have methods for configuration, fitting, and translation.

### For model specifications (before fitting):

| Method | Description | Arguments | Returns |
|--------|-------------|-----------|---------|
| `set_engine(spec, engine)` | Choose computational engine | `engine`, `...` (engine-specific args) | Updated spec |
| `set_mode(spec, mode)` | Set prediction mode | `"regression"` or `"classification"` | Updated spec |
| `set_args(spec, ...)` | Update model arguments | Named arguments to update | Updated spec |
| `translate(spec)` | Show underlying model call | `engine` (optional) | Printed translation |
| `fit(spec, formula, data)` | Fit model to data | `formula`, `data`, or `x`/`y` | Fitted model |
| `fit_xy(spec, x, y)` | Fit with matrix interface | `x` (predictors), `y` (outcome) | Fitted model |

```{r parsnip-spec-methods, eval=FALSE}
# Inspect what the spec will do
rf_spec <- rand_forest(trees = 500, mtry = 3) %>%
  set_engine("ranger") %>%
  set_mode("regression")

translate(rf_spec)  # Shows: ranger::ranger(num.trees = 500, mtry = 3, ...)

# Update arguments
rf_spec <- rf_spec %>%
  set_args(trees = 1000)

# Fit directly (without workflow)
rf_fit <- rf_spec %>%
  fit(Sale_Price ~ ., data = train)
```

### For fitted models (after fitting):

| Method | Description | Returns |
|--------|-------------|---------|
| `predict(fit, new_data)` | Generate predictions | Tibble with `.pred` column |
| `predict(fit, new_data, type)` | Predictions of specific type | Tibble (type-dependent columns) |
| `extract_fit_engine(fit)` | Get underlying fitted model | Engine-specific object |
| `tidy(fit)` | Extract model coefficients/info | Tibble of parameters |
| `glance(fit)` | Model-level summaries | Tibble of fit statistics |
| `augment(fit, new_data)` | Add predictions to data | Tibble with original + predictions |

```{r parsnip-fitted-methods, eval=FALSE}
# Different prediction types
predict(rf_fit, new_data = test)                    # Point predictions
predict(rf_fit, new_data = test, type = "conf_int") # Confidence intervals
predict(rf_fit, new_data = test, type = "raw")      # Engine-specific format

# For classification:
predict(logit_fit, test, type = "class")  # Predicted classes
predict(logit_fit, test, type = "prob")   # Class probabilities

# Extract underlying model for package-specific functions
ranger_obj <- extract_fit_engine(rf_fit)
ranger_obj$variable.importance  # Use ranger-specific features

# Tidy model output
tidy(lm_fit)     # Coefficients table
glance(lm_fit)   # R-squared, AIC, etc.
```

### Available prediction types by mode:

**Regression:**
- `type = "numeric"` (default): Predicted numeric values
- `type = "conf_int"`: Confidence/prediction intervals
- `type = "pred_int"`: Prediction intervals
- `type = "raw"`: Engine-specific format

**Classification:**
- `type = "class"` (default): Predicted class labels
- `type = "prob"`: Class probabilities
- `type = "conf_int"`: Confidence intervals for probabilities
- `type = "raw"`: Engine-specific format

---

# 4. Creating Workflows

Workflows combine preprocessing (recipe) and model specification.

## Basic Workflow

```{r basic-workflow}
# Create simple recipe (without outcome transformation for prediction compatibility)
simple_recipe <- recipe(Sale_Price ~ Gr_Liv_Area + Year_Built + Lot_Area,
  data = ames_train
) %>%
  step_normalize(all_numeric_predictors())

# Create workflow
lm_workflow <- workflow() %>%
  add_recipe(simple_recipe) %>%
  add_model(lm_spec)

# View workflow
lm_workflow
```

## Fit Workflow

```{r fit-workflow}
# Fit the workflow
lm_fit <- lm_workflow %>%
  fit(data = ames_train)

# Extract the fitted model
lm_fit %>%
  extract_fit_parsnip() %>%
  tidy()

# Make predictions
predictions <- lm_fit %>%
  predict(ames_test)

head(predictions)
```

## Multiple Workflows

```{r multiple-workflows}
# Create workflow set with different models
workflow_set <- workflow_set(
  preproc = list(simple = simple_recipe),
  models = list(
    lm = lm_spec,
    rf = rf_spec,
    tree = tree_spec
  ),
  cross = TRUE
)

workflow_set
```

## Key Methods for Workflow Objects

Workflows bundle preprocessing and modeling, with methods for construction, fitting, and extraction.

### Building workflows:

| Method | Description | Arguments | Returns |
|--------|-------------|-----------|---------|
| `workflow()` | Create empty workflow | None | Empty workflow |
| `add_recipe(wf, recipe)` | Add preprocessing recipe | `recipe` object | Updated workflow |
| `add_model(wf, spec)` | Add model specification | `spec` (parsnip model) | Updated workflow |
| `add_formula(wf, formula)` | Add formula (no recipe) | R formula | Updated workflow |
| `add_variables(wf, outcomes, predictors)` | Add variables directly | Variable selectors | Updated workflow |
| `remove_recipe(wf)` | Remove recipe | None | Updated workflow |
| `remove_model(wf)` | Remove model | None | Updated workflow |
| `update_recipe(wf, recipe)` | Replace recipe | New `recipe` | Updated workflow |
| `update_model(wf, spec)` | Replace model | New `spec` | Updated workflow |

```{r workflow-building, eval=FALSE}
# Build workflow step by step
wf <- workflow() %>%
  add_recipe(my_recipe) %>%
  add_model(rf_spec)

# Update components
wf <- wf %>%
  update_model(xgb_spec)  # Switch to different model

# Remove and re-add
wf <- wf %>%
  remove_recipe() %>%
  add_formula(y ~ x1 + x2)  # Use formula instead
```

### Fitting and using workflows:

| Method | Description | Arguments | Returns |
|--------|-------------|-----------|---------|
| `fit(wf, data)` | Fit workflow to data | Training data | Fitted workflow |
| `predict(fitted_wf, new_data)` | Generate predictions | New data, `type` | Tibble with predictions |
| `last_fit(wf, split)` | Fit and evaluate on split | `initial_split` object | Special results object |
| `fit_resamples(wf, resamples)` | Fit on resamples | Cross-validation folds | Resampling results |
| `tune_grid(wf, resamples, grid)` | Tune hyperparameters | Resamples, parameter grid | Tuning results |

```{r workflow-fitting, eval=FALSE}
# Basic fitting
fitted_wf <- wf %>%
  fit(data = train)

# Predict
predictions <- fitted_wf %>%
  predict(new_data = test)

# Fit and immediately evaluate on test set
final_results <- wf %>%
  last_fit(split = train_test_split)

final_results %>% collect_metrics()  # Test set metrics
```

### Extracting from fitted workflows:

| Method | Description | Returns |
|--------|-------------|---------|
| `extract_recipe(fitted_wf)` | Get fitted recipe | Prepped recipe |
| `extract_fit_parsnip(fitted_wf)` | Get parsnip model fit | Fitted parsnip object |
| `extract_fit_engine(fitted_wf)` | Get underlying model | Engine-specific object |
| `extract_spec_parsnip(fitted_wf)` | Get model specification | Parsnip spec |
| `extract_preprocessor(fitted_wf)` | Get preprocessor (recipe/formula) | Recipe or formula |
| `extract_mold(fitted_wf)` | Get processed training data | Mold object |

```{r workflow-extraction, eval=FALSE}
# Extract components
fitted_recipe <- extract_recipe(fitted_wf)
fitted_model <- extract_fit_parsnip(fitted_wf)
engine_model <- extract_fit_engine(fitted_wf)

# Use extracted components
tidy(fitted_model)  # Model coefficients
bake(fitted_recipe, new_data = new_data)  # Apply preprocessing

# Access underlying model for package-specific features
engine_model$variable.importance  # If using ranger
```

### Workflow sets (multiple workflows):

| Method | Description | Arguments | Returns |
|--------|-------------|-----------|---------|
| `workflow_set(preproc, models)` | Create multiple workflows | Lists of recipes/models | Workflow set |
| `workflow_map(wf_set, fn)` | Apply function to all | Function name, resamples | Updated set |
| `rank_results(wf_set)` | Rank by performance | `rank_metric` | Ranked tibble |
| `autoplot(wf_set)` | Visualize results | `metric` | ggplot object |

```{r workflow-set-methods, eval=FALSE}
# Create and fit multiple workflows
wf_set <- workflow_set(
  preproc = list(basic = recipe1, engineered = recipe2),
  models = list(rf = rf_spec, xgb = xgb_spec)
) %>%
  workflow_map(fn = "fit_resamples", resamples = folds)

# Compare results
wf_set %>%
  rank_results(rank_metric = "rmse")

# Visualize
wf_set %>%
  autoplot(metric = "rmse")
```

---

# 5. Hyperparameter Tuning

## Define Tunable Parameters

```{r tuning-setup}
# Create a model with tunable parameters
rf_tune_spec <- rand_forest(
  trees = 1000,
  mtry = tune(), # Number of predictors to sample
  min_n = tune() # Minimum observations in node
) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Create workflow
rf_tune_workflow <- workflow() %>%
  add_recipe(simple_recipe) %>%
  add_model(rf_tune_spec)
```

## Grid Search

```{r grid-search, eval=FALSE}
# Create regular grid
rf_grid <- grid_regular(
  mtry(range = c(1, 5)),
  min_n(range = c(2, 10)),
  levels = 5
)

# Tune with cross-validation
rf_tune_results <- rf_tune_workflow %>%
  tune_grid(
    resamples = ames_folds,
    grid = rf_grid,
    metrics = metric_set(rmse, rsq)
  )

# View results
rf_tune_results %>%
  collect_metrics() %>%
  arrange(mean)

# Select best parameters
best_rf <- rf_tune_results %>%
  select_best(metric = "rmse")

# Finalize workflow with best parameters
final_rf_workflow <- rf_tune_workflow %>%
  finalize_workflow(best_rf)

# Fit final model
final_rf_fit <- final_rf_workflow %>%
  last_fit(ames_split)

# Get test set performance
final_rf_fit %>%
  collect_metrics()
```

## Random Search

```{r random-search, eval=FALSE}
# Create random grid
rf_random_grid <- grid_random(
  mtry(range = c(1, 10)),
  min_n(range = c(2, 20)),
  size = 20 # Number of combinations
)

# Tune with random search
rf_random_results <- rf_tune_workflow %>%
  tune_grid(
    resamples = ames_folds,
    grid = rf_random_grid
  )
```

## Space-Filling Designs (Irregular Grids)

For more efficient parameter space exploration, use **space-filling designs** like Latin hypercube:

```{r space-filling, eval=FALSE}
# Latin hypercube sampling - better coverage of parameter space
latin_grid <- grid_latin_hypercube(
  penalty(range = c(-5, 0)),
  mixture(range = c(0, 1)),
  size = 20 # Number of parameter combinations
)

# Advantages over regular grid:
# - Better coverage with fewer combinations
# - Avoids clustering of points
# - More efficient for expensive models
# - Recommended for irregular/non-linear parameter relationships

# Example: Compare grid sizes
regular_grid <- grid_regular(penalty(), mixture(), levels = 5) # 5x5 = 25 points
latin_grid <- grid_latin_hypercube(penalty(), mixture(), size = 20) # 20 points, better coverage
```

## Bayesian Optimization

```{r bayesian-opt, eval=FALSE}
library(finetune)

# Bayesian optimization
rf_bayes_results <- rf_tune_workflow %>%
  tune_bayes(
    resamples = ames_folds,
    initial = 5, # Initial random parameter sets
    iter = 20, # Additional iterations
    metrics = metric_set(rmse)
  )

# View best results
rf_bayes_results %>%
  show_best(metric = "rmse")
```

## Key Methods for Tuning Results

After tuning, these methods help you explore results, select best parameters, and finalize models.

### Exploring tuning results:

| Method | Description | Arguments | Returns |
|--------|-------------|-----------|---------|
| `collect_metrics(results)` | Get all metrics across parameters | `summarize = TRUE` | Tibble of metrics |
| `show_best(results, metric)` | Show top parameter sets | `metric`, `n` | Tibble (best n sets) |
| `autoplot(results)` | Visualize tuning results | `metric` | ggplot object |
| `collect_predictions(results)` | Get predictions from resamples | `summarize = FALSE` | Tibble of predictions |
| `conf_mat_resampled(results)` | Confusion matrix (classification) | None | Summary object |

```{r tuning-exploration, eval=FALSE}
# Get all metrics
all_metrics <- tune_results %>%
  collect_metrics()

# See best 5 parameter combinations
tune_results %>%
  show_best(metric = "rmse", n = 5)

# Visualize performance
tune_results %>%
  autoplot(metric = "rmse")

# Get all predictions
all_preds <- tune_results %>%
  collect_predictions(summarize = FALSE)
```

### Selecting best parameters:

| Method | Description | Arguments | Returns |
|--------|-------------|-----------|---------|
| `select_best(results, metric)` | Choose best by metric | `metric` | Tibble (1 row) |
| `select_by_pct_loss(results)` | Select within % of best | `metric`, `limit` | Tibble (1 row) |
| `select_by_one_std_err(results)` | One SE rule selection | `metric` | Tibble (1 row) |

```{r parameter-selection, eval=FALSE}
# Select absolute best
best_params <- tune_results %>%
  select_best(metric = "rmse")

# Select simpler model within 2% of best performance
simple_params <- tune_results %>%
  select_by_pct_loss(metric = "rmse", limit = 2)

# One standard error rule (simpler model)
simple_params <- tune_results %>%
  select_by_one_std_err(metric = "rmse", desc(trees))  # Prefer fewer trees
```

### Finalizing workflows:

| Method | Description | Arguments | Returns |
|--------|-------------|-----------|---------|
| `finalize_workflow(wf, params)` | Update workflow with best params | Workflow, parameter tibble | Updated workflow |
| `finalize_model(spec, params)` | Update model spec with params | Model spec, parameters | Updated spec |
| `finalize_recipe(recipe, params)` | Update recipe with params | Recipe, parameters | Updated recipe |

```{r finalize-workflow, eval=FALSE}
# Get best parameters
best <- tune_results %>%
  select_best("rmse")

# Finalize workflow
final_wf <- tune_workflow %>%
  finalize_workflow(best)

# Fit on all training data
final_fit <- final_wf %>%
  fit(data = train)

# Or use last_fit for automatic train/test evaluation
final_results <- tune_workflow %>%
  finalize_workflow(best) %>%
  last_fit(split = train_test_split)

final_results %>% collect_metrics()  # Test set performance
```

### Special methods for `last_fit()` results:

| Method | Description | Returns |
|--------|-------------|---------|
| `collect_metrics(last_fit)` | Test set metrics | Tibble of metrics |
| `collect_predictions(last_fit)` | Test set predictions | Tibble with predictions |
| `extract_workflow(last_fit)` | Get fitted workflow | Fitted workflow |

```{r last-fit-methods, eval=FALSE}
# Evaluate on test set
test_metrics <- final_results %>%
  collect_metrics()

# Get test predictions
test_preds <- final_results %>%
  collect_predictions()

# Extract for deployment
deployment_wf <- final_results %>%
  extract_workflow()

# Save for later
saveRDS(deployment_wf, "final_model.rds")
```

---

# 6. Model Evaluation with yardstick

## Regression Metrics

```{r regression-metrics, eval=FALSE}
# Get predictions with actual values
results <- lm_fit %>%
  predict(ames_test) %>%
  bind_cols(ames_test %>% select(Sale_Price))

# Calculate metrics
metrics_regression <- metric_set(rmse, rsq, mae, mape)

results %>%
  metrics_regression(truth = Sale_Price, estimate = .pred)

# Individual metrics
results %>%
  rmse(truth = Sale_Price, estimate = .pred)

results %>%
  rsq(truth = Sale_Price, estimate = .pred)
```

## Classification Metrics

```{r classification-metrics, eval=FALSE}
# For binary classification
# Assume we have predictions and true classes
classification_results <- tibble(
  truth = factor(c("Yes", "No", "Yes", "Yes", "No")),
  estimate = factor(c("Yes", "No", "No", "Yes", "No")),
  .pred_Yes = c(0.9, 0.3, 0.4, 0.8, 0.2)
)

# Confusion matrix
classification_results %>%
  conf_mat(truth = truth, estimate = estimate)

# Multiple metrics
metrics_classification <- metric_set(accuracy, sensitivity, specificity, f_meas)

classification_results %>%
  metrics_classification(truth = truth, estimate = estimate)

# ROC curve
classification_results %>%
  roc_curve(truth = truth, .pred_Yes) %>%
  autoplot()

# AUC
classification_results %>%
  roc_auc(truth = truth, .pred_Yes)
```

## Key Methods for yardstick Metrics

Yardstick provides consistent functions for calculating model performance metrics.

### Using individual metrics:

All metric functions follow the same pattern:
```r
metric_function(data, truth, estimate, ...)
```

**Common regression metrics:**

| Function | Metric | Range | Better |
|----------|--------|-------|--------|
| `rmse()` | Root Mean Squared Error | [0, ∞] | Lower |
| `mae()` | Mean Absolute Error | [0, ∞] | Lower |
| `rsq()` | R-squared | [0, 1] | Higher |
| `mape()` | Mean Absolute % Error | [0, ∞] | Lower |
| `smape()` | Symmetric MAPE | [0, 200] | Lower |
| `msd()` | Mean Signed Deviation | (-∞, ∞) | Close to 0 |
| `ccc()` | Concordance Correlation | [-1, 1] | Higher |
| `rpd()` | Ratio of Performance to Deviation | [0, ∞] | Higher |
| `rpiq()` | Ratio of Performance to IQR | [0, ∞] | Higher |
| `huber_loss()` | Huber Loss (robust) | [0, ∞] | Lower |

**Common classification metrics:**

| Function | Metric | Range | Better |
|----------|--------|-------|--------|
| `accuracy()` | Overall accuracy | [0, 1] | Higher |
| `bal_accuracy()` | Balanced accuracy | [0, 1] | Higher |
| `sens()` / `sensitivity()` | True positive rate | [0, 1] | Higher |
| `spec()` / `specificity()` | True negative rate | [0, 1] | Higher |
| `precision()` / `ppv()` | Positive predictive value | [0, 1] | Higher |
| `recall()` | Same as sensitivity | [0, 1] | Higher |
| `f_meas()` | F1-score | [0, 1] | Higher |
| `kap()` | Cohen's Kappa | [-1, 1] | Higher |
| `mcc()` | Matthews Correlation | [-1, 1] | Higher |
| `roc_auc()` | Area under ROC curve | [0, 1] | Higher |
| `pr_auc()` | Area under PR curve | [0, 1] | Higher |

```{r yardstick-individual, eval=FALSE}
# Single metric
results %>%
  rmse(truth = actual, estimate = predicted)

# With grouped data
results %>%
  group_by(model_type) %>%
  rmse(truth = actual, estimate = predicted)
```

### Creating metric sets:

| Function | Description | Arguments | Returns |
|----------|-------------|-----------|---------|
| `metric_set(...)` | Combine multiple metrics | Metric functions | Metric set function |
| `metrics(data, truth, estimate)` | Default metrics | Data with truth/estimate | Tibble of metrics |

```{r metric-sets, eval=FALSE}
# Create custom metric set
my_metrics <- metric_set(rmse, rsq, mae, mape)

# Apply all at once
results %>%
  my_metrics(truth = actual, estimate = predicted)

# Default metrics (quick evaluation)
results %>%
  metrics(truth = actual, estimate = predicted)
```

### Confusion matrix methods:

| Method | Description | Returns |
|--------|-------------|---------|
| `conf_mat(data, truth, estimate)` | Create confusion matrix | conf_mat object |
| `summary(conf_mat)` | All classification metrics | Tibble of metrics |
| `autoplot(conf_mat)` | Visualize confusion matrix | ggplot object |
| `tidy(conf_mat)` | Convert to tibble | Tibble |

```{r conf-mat-methods, eval=FALSE}
# Create confusion matrix
cm <- classification_results %>%
  conf_mat(truth = truth, estimate = estimate)

# Get all metrics at once
summary(cm)

# Visualize
autoplot(cm, type = "heatmap")
autoplot(cm, type = "mosaic")

# Extract as data
tidy(cm)
```

### ROC and PR curves:

| Function | Description | Returns |
|----------|-------------|---------|
| `roc_curve(data, truth, ...)` | Calculate ROC curve points | Tibble of sensitivity/specificity |
| `pr_curve(data, truth, ...)` | Calculate PR curve points | Tibble of precision/recall |
| `gain_curve(data, truth, ...)` | Cumulative gain curve | Tibble |
| `lift_curve(data, truth, ...)` | Lift curve | Tibble |
| `autoplot(curve)` | Visualize curve | ggplot object |

```{r curve-methods, eval=FALSE}
# ROC curve
roc_data <- classification_results %>%
  roc_curve(truth = truth, .pred_Yes)

autoplot(roc_data)

# PR curve
pr_data <- classification_results %>%
  pr_curve(truth = truth, .pred_Yes)

autoplot(pr_data)

# Compare multiple models
bind_rows(
  roc_curve(results1, truth, .pred_Yes) %>% mutate(model = "Model 1"),
  roc_curve(results2, truth, .pred_Yes) %>% mutate(model = "Model 2")
) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path() +
  geom_abline(linetype = "dashed")
```

### Grouped and multi-class metrics:

```{r grouped-metrics, eval=FALSE}
# Metrics by group
results %>%
  group_by(model_type) %>%
  rmse(truth = actual, estimate = predicted)

# Multi-class classification (macro/micro averaging)
multiclass_results %>%
  accuracy(truth = class, estimate = predicted_class)

multiclass_results %>%
  roc_auc(
    truth = class,
    .pred_A, .pred_B, .pred_C,  # Probability columns
    estimator = "macro"  # or "micro", "macro_weighted"
  )
```

---

# 7. Complete Example: End-to-End Workflow

```{r complete-example, eval=FALSE}
# Load data
data(ames)
set.seed(123)

# 1. Split data
ames_split <- initial_split(ames, prop = 0.75, strata = Sale_Price)
ames_train <- training(ames_split)
ames_test <- testing(ames_split)

# Create cross-validation folds
ames_folds <- vfold_cv(ames_train, v = 5, strata = Sale_Price)

# 2. Create recipe
ames_recipe <- recipe(
  Sale_Price ~ Lot_Area + Neighborhood + Gr_Liv_Area +
    Year_Built + Bldg_Type + Overall_Qual,
  data = ames_train
) %>%
  step_log(Sale_Price, Gr_Liv_Area, Lot_Area) %>%
  step_other(Neighborhood, threshold = 0.05) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

# 3. Specify models with tuning
xgb_spec <- boost_tree(
  trees = 1000,
  tree_depth = tune(),
  learn_rate = tune(),
  mtry = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# 4. Create workflow
xgb_workflow <- workflow() %>%
  add_recipe(ames_recipe) %>%
  add_model(xgb_spec)

# 5. Define tuning grid
xgb_grid <- grid_latin_hypercube(
  tree_depth(range = c(3, 10)),
  learn_rate(range = c(-3, -0.5)),
  finalize(mtry(), ames_train %>% select(-Sale_Price)),
  size = 20
)

# 6. Tune model
xgb_tune <- xgb_workflow %>%
  tune_grid(
    resamples = ames_folds,
    grid = xgb_grid,
    control = control_grid(save_pred = TRUE),
    metrics = metric_set(rmse, rsq, mae)
  )

# 7. Evaluate tuning results
xgb_tune %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  arrange(mean)

# Visualize tuning results
xgb_tune %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = learn_rate, y = mean, color = factor(tree_depth))) +
  geom_point() +
  geom_line() +
  scale_x_log10() +
  labs(
    title = "Tuning Results: RMSE by Learning Rate and Tree Depth",
    y = "Mean RMSE",
    color = "Tree Depth"
  )

# 8. Select best model
best_xgb <- xgb_tune %>%
  select_best(metric = "rmse")

# 9. Finalize workflow
final_xgb <- xgb_workflow %>%
  finalize_workflow(best_xgb)

# 10. Fit on all training data and evaluate on test set
final_fit <- final_xgb %>%
  last_fit(ames_split)

# 11. Get test set performance
final_fit %>%
  collect_metrics()

# 12. Get predictions
test_predictions <- final_fit %>%
  collect_predictions()

test_predictions %>%
  ggplot(aes(x = Sale_Price, y = .pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(color = "red", linetype = "dashed") +
  labs(
    title = "Predicted vs Actual Sale Price",
    x = "Actual (log10)",
    y = "Predicted (log10)"
  )

# 13. Fit final model on all data for deployment
final_model <- final_xgb %>%
  fit(ames)

# Save model for later use
# saveRDS(final_model, "ames_price_model.rds")
```

---

# 8. Advanced Topics

## Custom Metrics

```{r custom-metrics, eval=FALSE}
# Create custom metric
mean_absolute_percentage_error <- function(data, truth, estimate, ...) {
  mape_impl <- function(truth, estimate) {
    mean(abs((truth - estimate) / truth)) * 100
  }

  metric_summarizer(
    metric_nm = "mape",
    metric_fn = mape_impl,
    data = data,
    truth = !!rlang::enquo(truth),
    estimate = !!rlang::enquo(estimate),
    ...
  )
}

# Use custom metric
custom_metrics <- metric_set(rmse, rsq, mean_absolute_percentage_error)
```

## Model Stacking

```{r stacking, eval=FALSE}
library(stacks)

# Create multiple tuned models
# ... (tune several models as shown above)

# Initialize stack
model_stack <- stacks() %>%
  add_candidates(rf_tune_results) %>%
  add_candidates(xgb_tune) %>%
  add_candidates(lm_tune_results)

# Blend predictions
blended_stack <- model_stack %>%
  blend_predictions()

# Fit stack members
fitted_stack <- blended_stack %>%
  fit_members()

# Predict with stack
stack_pred <- fitted_stack %>%
  predict(ames_test)
```

## Feature Importance

```{r feature-importance, eval=FALSE}
library(vip)

# For models with built-in importance
lm_fit %>%
  extract_fit_parsnip() %>%
  vip()

# Model-agnostic importance with DALEX
library(DALEXtra)

explainer <- explain_tidymodels(
  final_model,
  data = ames_train %>% select(-Sale_Price),
  y = ames_train$Sale_Price,
  label = "XGBoost"
)

# Variable importance
vi <- model_parts(explainer)
plot(vi)

# Partial dependence plots
pdp <- model_profile(explainer, variables = "Gr_Liv_Area")
plot(pdp)
```

## Parallel Processing

```{r parallel, eval=FALSE}
library(doParallel)

# Setup parallel backend
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

# Tuning will now use parallel processing
tune_results <- workflow %>%
  tune_grid(
    resamples = folds,
    grid = tuning_grid,
    control = control_grid(parallel_over = "resamples")
  )

# Stop cluster when done
stopCluster(cl)
```

## Comparing Modeling Frameworks: tidymodels vs caret vs Base R

Understanding the differences between R's modeling frameworks helps you choose the right tool for your needs.

### Overview Comparison

| Feature | tidymodels | caret | Base R |
|---------|-----------|-------|--------|
| **Philosophy** | Modular, tidyverse-aligned | All-in-one wrapper | Direct model access |
| **Learning Curve** | Moderate | Moderate | Varies by package |
| **Code Style** | Pipe-friendly (`%>%`) | Function-based | Function-based |
| **Consistency** | High (unified interface) | High (unified wrapper) | Low (package-specific) |
| **Preprocessing** | recipes package | preProcess, formula | Manual or package-specific |
| **Model Specification** | parsnip (declarative) | train() method argument | Package-specific functions |
| **Active Development** | Very active | Maintenance mode* | Package-specific |
| **Extensibility** | Highly modular | Limited | Unlimited |

*Note: caret is in maintenance mode as Max Kuhn (creator) moved to tidymodels

### Detailed Comparison with Examples

#### 1. Data Splitting

```{r comparison-splitting, eval=FALSE}
# tidymodels
library(tidymodels)
set.seed(123)
split <- initial_split(mtcars, prop = 0.75, strata = mpg)
train_data <- training(split)
test_data <- testing(split)
cv_folds <- vfold_cv(train_data, v = 10)

# caret
library(caret)
set.seed(123)
train_index <- createDataPartition(mtcars$mpg, p = 0.75, list = FALSE)
train_data <- mtcars[train_index, ]
test_data <- mtcars[-train_index, ]
cv_folds <- createFolds(train_data$mpg, k = 10)

# Base R
set.seed(123)
train_index <- sample(1:nrow(mtcars), size = 0.75 * nrow(mtcars))
train_data <- mtcars[train_index, ]
test_data <- mtcars[-train_index, ]
# CV folds require manual implementation
```

#### 2. Preprocessing and Feature Engineering

```{r comparison-preprocessing, eval=FALSE}
# tidymodels - recipes
recipe_tm <- recipe(mpg ~ ., data = train_data) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_corr(all_numeric_predictors(), threshold = 0.9)

# Apply to new data
prepped_recipe <- prep(recipe_tm, training = train_data)
train_processed <- bake(prepped_recipe, new_data = train_data)
test_processed <- bake(prepped_recipe, new_data = test_data)

# caret - preProcess
preproc_caret <- preProcess(
  train_data[, -1], # Exclude outcome
  method = c("center", "scale", "zv", "corr")
)
train_processed <- predict(preproc_caret, train_data)
test_processed <- predict(preproc_caret, test_data)

# Base R - manual
train_means <- colMeans(train_data[, sapply(train_data, is.numeric)])
train_sds <- apply(train_data[, sapply(train_data, is.numeric)], 2, sd)
train_processed <- scale(train_data[, -1], center = train_means[-1], scale = train_sds[-1])
test_processed <- scale(test_data[, -1], center = train_means[-1], scale = train_sds[-1])
# Dummy coding, correlation removal, etc. require manual implementation
```

#### 3. Model Training

```{r comparison-training, eval=FALSE}
# tidymodels - workflow approach
rf_spec <- rand_forest(trees = 500, mtry = 3) %>%
  set_engine("ranger") %>%
  set_mode("regression")

wf <- workflow() %>%
  add_recipe(recipe_tm) %>%
  add_model(rf_spec)

fitted_model <- wf %>%
  fit(data = train_data)

# caret - unified interface
fitted_model_caret <- train(
  mpg ~ .,
  data = train_data,
  method = "ranger",
  preProcess = c("center", "scale"),
  trControl = trainControl(method = "cv", number = 10),
  tuneGrid = expand.grid(
    mtry = 3,
    splitrule = "variance",
    min.node.size = 5
  ),
  num.trees = 500
)

# Base R - direct package calls
library(ranger)
fitted_model_base <- ranger(
  mpg ~ .,
  data = train_processed, # Must preprocess first
  num.trees = 500,
  mtry = 3
)
```

#### 4. Hyperparameter Tuning

```{r comparison-tuning, eval=FALSE}
# tidymodels - declarative tuning
rf_tune_spec <- rand_forest(
  trees = 500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("regression")

wf_tune <- workflow() %>%
  add_recipe(recipe_tm) %>%
  add_model(rf_tune_spec)

tune_results <- wf_tune %>%
  tune_grid(
    resamples = cv_folds,
    grid = 20,
    metrics = metric_set(rmse, rsq)
  )

best_params <- select_best(tune_results, "rmse")

# caret - grid search built-in
tune_grid_caret <- expand.grid(
  mtry = c(2, 3, 4, 5),
  splitrule = "variance",
  min.node.size = c(5, 10, 15)
)

tuned_model_caret <- train(
  mpg ~ .,
  data = train_data,
  method = "ranger",
  trControl = trainControl(method = "cv", number = 10),
  tuneGrid = tune_grid_caret,
  num.trees = 500
)

# Base R - manual grid search
param_grid <- expand.grid(
  mtry = c(2, 3, 4, 5),
  min.node.size = c(5, 10, 15)
)

cv_results <- data.frame()
for (i in 1:nrow(param_grid)) {
  # Implement CV loop manually
  fold_rmse <- numeric(10)
  for (fold in 1:10) {
    # Split data into train/validation
    # Train model with params
    # Calculate RMSE
  }
  # Store results
}
# Select best parameters
```

#### 5. Predictions and Evaluation

```{r comparison-predictions, eval=FALSE}
# tidymodels
predictions_tm <- fitted_model %>%
  predict(test_data) %>%
  bind_cols(test_data)

metrics_tm <- predictions_tm %>%
  metrics(truth = mpg, estimate = .pred)

# caret
predictions_caret <- predict(fitted_model_caret, newdata = test_data)
metrics_caret <- postResample(pred = predictions_caret, obs = test_data$mpg)

# Base R
predictions_base <- predict(fitted_model_base, data = test_data)$predictions
rmse_base <- sqrt(mean((test_data$mpg - predictions_base)^2))
rsq_base <- cor(test_data$mpg, predictions_base)^2
```

### Pros and Cons

#### tidymodels

**Pros:**
- **Modern design**: Built with tidyverse principles, pipe-friendly
- **Modularity**: Each package has a focused purpose, easy to extend
- **Reproducibility**: Recipes ensure consistent preprocessing across train/test
- **Active development**: New features, models, and improvements regularly added
- **Better documentation**: Comprehensive website and "Tidy Modeling with R" book
- **Type safety**: Strong typing and error checking prevent common mistakes
- **Workflow objects**: Preprocessing + model bundled together, easier deployment
- **Advanced features**: Model stacking, Bayesian optimization, workflow sets
- **Consistent interface**: Same syntax across 100+ models

**Cons:**
- **Newer ecosystem**: Less mature than caret, some rough edges
- **Steeper learning curve initially**: Need to understand multiple packages
- **More verbose**: Requires more code than caret for simple tasks
- **Performance overhead**: Recipes re-apply preprocessing in each resample (correctness over speed)
- **Training speed**: Can be significantly slower than caret for tuning (48× in one benchmark)
- **Breaking changes**: API still evolving (though stabilizing)

**Best for:**
- Production pipelines requiring reproducibility
- Complex preprocessing workflows
- Projects using tidyverse extensively
- Teams wanting consistent, maintainable code
- Advanced modeling techniques (ensembles, stacking)

**Important preprocessing philosophy**:
Tidymodels prioritizes **correctness over speed**. Recipes ensure preprocessing statistics are computed only on training folds and applied to validation folds within each resample, preventing data leakage. This is more rigorous but slower than caret's approach.

#### caret

**Pros:**
- **Mature and stable**: 15+ years of development, well-tested
- **Comprehensive**: 238+ models with unified interface
- **Simple syntax**: Less verbose for basic tasks
- **Extensive documentation**: Many tutorials and examples available
- **Feature engineering**: Built-in functions for preprocessing
- **Variable importance**: Standardized across models
- **Large community**: Lots of Stack Overflow answers and blog posts

**Cons:**
- **Maintenance mode**: No major new features (creator moved to tidymodels)
- **Less flexible preprocessing**: preProcess limited compared to recipes
- **Harder to extend**: Adding new models or methods requires package updates
- **Not tidyverse-native**: Doesn't integrate well with dplyr, ggplot2, etc.
- **Formula interface limitations**: Can be clunky for complex feature engineering
- **Inconsistent behavior**: Some models have unique quirks despite wrapper
- **Memory intensive**: Can be slow with large datasets

**Best for:**
- Quick prototyping and experimentation
- Legacy projects already using caret
- Users familiar with traditional R syntax
- Simple modeling tasks without complex preprocessing
- When you need one of caret's specialized features

#### Base R

**Pros:**
- **Full control**: Direct access to all model parameters and internals
- **No dependencies**: Lightweight, no extra packages needed (for base models)
- **Performance**: No abstraction overhead, can be fastest option
- **Flexibility**: Can implement any custom logic
- **Stability**: Core functions rarely change
- **Learning**: Understanding base functions teaches fundamentals

**Cons:**
- **Inconsistent APIs**: Each package has different syntax and conventions
- **Manual work**: Must implement train/test splits, CV, preprocessing manually
- **Error-prone**: Easy to make data leakage mistakes
- **More code**: Requires significantly more lines for complete workflows
- **Limited scalability**: Hard to maintain as projects grow
- **Harder to reproduce**: More manual steps mean more places for errors

**Best for:**
- Learning fundamentals of machine learning
- Custom algorithms or research
- Simple models without preprocessing
- Performance-critical applications
- When you need package-specific features not exposed by wrappers

### Migration Guide

#### From caret to tidymodels

```{r migration-guide, eval=FALSE}
# caret code
library(caret)
model <- train(
  mpg ~ .,
  data = train_data,
  method = "rf",
  preProcess = c("center", "scale"),
  trControl = trainControl(method = "cv", number = 10)
)

# Equivalent tidymodels code
library(tidymodels)

# Split data
split <- initial_split(mtcars, prop = 0.75)
train_data <- training(split)

# Create recipe (preprocessing)
rec <- recipe(mpg ~ ., data = train_data) %>%
  step_normalize(all_numeric_predictors())

# Create model spec
rf_spec <- rand_forest() %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Create workflow
wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_spec)

# Fit with cross-validation
cv_folds <- vfold_cv(train_data, v = 10)
cv_results <- wf %>%
  fit_resamples(resamples = cv_folds)

# Final fit
final_fit <- wf %>%
  fit(train_data)
```

#### From Base R to tidymodels

```{r base-to-tidymodels, eval=FALSE}
# Base R code
library(randomForest)
train_idx <- sample(1:nrow(mtcars), 0.75 * nrow(mtcars))
train <- mtcars[train_idx, ]
test <- mtcars[-train_idx, ]
train_scaled <- scale(train[, -1])
model <- randomForest(mpg ~ ., data = as.data.frame(train_scaled))
predictions <- predict(model, newdata = scale(test[, -1]))

# tidymodels equivalent
library(tidymodels)
split <- initial_split(mtcars, prop = 0.75)
rec <- recipe(mpg ~ ., data = training(split)) %>%
  step_normalize(all_numeric_predictors())
rf_spec <- rand_forest() %>%
  set_engine("randomForest") %>%
  set_mode("regression")
wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_spec)
fit <- wf %>% fit(training(split))
predictions <- fit %>% predict(testing(split))
```

### Recommendation

**Choose tidymodels if:**
- Starting a new project in 2024+
- Need reproducible, production-ready pipelines
- Working with complex preprocessing
- Want modern R programming practices
- Plan to use advanced techniques (stacking, tuning)

**Choose caret if:**
- Maintaining existing caret code
- Need quick prototyping with minimal setup
- Have simple preprocessing needs
- Want maximum model variety with minimal code

**Choose Base R if:**
- Learning ML fundamentals
- Building custom algorithms
- Need maximum performance and control
- Have very simple modeling needs
- Working with specialized packages not wrapped by tidymodels/caret

**General guideline**: For new projects, **tidymodels** is the recommended choice as it's the future of modeling in R and provides the best balance of power, flexibility, and maintainability.

### Real-World Performance Comparison

A detailed comparison using the IBM HR Analytics dataset (employee churn prediction) revealed important practical differences:

#### Training Speed

**Critical Finding**: In a real-world benchmark, **caret was 48× faster** than tidymodels for hyperparameter tuning:
- tidymodels: 6+ hours for workflow_map() with 3,240 iterations
- caret: < 8 minutes for the same task

**Why the difference?**
- Recipes in tidymodels re-apply all preprocessing steps within each resample
- This ensures perfect reproducibility but adds computational overhead
- Caret caches preprocessing results more efficiently

**Mitigation strategies for tidymodels:**
1. Use smaller tuning grids initially
2. Leverage parallel processing with `doParallel`
3. Consider simpler recipes when speed is critical
4. Use `workflow_map()` with `control = control_grid(save_pred = FALSE)`

#### Model Performance

Surprisingly, in the benchmark study:
- **caret achieved higher F1-scores** on the test set
- This was attributed to caret's targeted optimization (explicitly optimizing for F1)
- tidymodels used default metrics which may not have prioritized F1

**Key lesson**: Always specify your optimization metric explicitly in tidymodels:
```r
tune_grid(
  resamples = folds,
  grid = tuning_grid,
  metrics = metric_set(f_meas, roc_auc, accuracy)  # Specify your priority metric
)
```

#### Workflow Differences

**tidymodels approach:**
```r
# Create workflow set for multiple models
workflow_set <- workflow_set(
  preproc = list(recipe1, recipe2),
  models = list(glmnet = glmnet_spec, xgb = xgb_spec),
  cross = TRUE
)

# Tune all at once
results <- workflow_set %>%
  workflow_map(fn = "tune_grid", resamples = folds)
```

**caret approach:**
```r
# Must create custom function to compare models
train_model <- function(method) {
  train(outcome ~ ., data = train_data, method = method,
        trControl = trainControl(...))
}

models <- lapply(c("glmnet", "xgbTree"), train_model)
```

#### Cross-Package Compatibility

For fair comparisons between frameworks:
```r
# Convert resamples between packages
caret_folds <- rsample2caret(tidymodels_folds)
tidymodels_folds <- caret2rsample(caret_folds)
```

### When Speed vs Features Matters

**Choose caret when:**
- Time-constrained exploratory analysis
- Need to quickly prototype many models
- Working with standard preprocessing workflows
- CPU resources are limited

**Choose tidymodels when:**
- Building production pipelines (reproducibility > speed)
- Complex, custom preprocessing required
- Need explicit workflow documentation
- Long-term maintainability matters more than initial speed
- Working in tidyverse-heavy projects

### Summary Table: Practical Considerations

| Consideration | tidymodels | caret |
|---------------|-----------|-------|
| **Initial setup time** | Longer (more code) | Faster (compact syntax) |
| **Tuning speed** | Slower (recipe overhead) | Faster (cached preprocessing) |
| **Reproducibility** | Excellent (explicit recipes) | Good (but less transparent) |
| **Debugging** | Easier (step-by-step) | Harder (black box) |
| **Memory usage** | Higher | Lower |
| **Parallelization** | Built-in support | Built-in support |
| **Custom metrics** | Easy to define | More complex |
| **Deployment** | Workflow objects are self-contained | Requires manual tracking |

---

# 9. Using btw to Learn More

The `btw` package is excellent for learning tidymodels functions interactively.

## Learn About Specific Functions

```{r btw-examples, eval=FALSE}
library(btw)

# Learn about recipes
btw(recipe, ames_train)

# Learn about specific preprocessing steps
btw(step_normalize, step_dummy, step_impute_median)

# Learn about model specification
btw(rand_forest, ames_train, "Show me examples of random forest regression")

# Learn about tuning
btw(tune_grid, ames_folds, "How do I tune hyperparameters?")

# Compare related functions
btw(
  initial_split, vfold_cv, bootstraps,
  "What's the difference between these resampling methods?"
)

# Learn about metrics
btw(rmse, rsq, mae, "Explain these regression metrics")

# Workflow creation
btw(workflow, "Show me how to create and use workflows")
```

## Get Package Overview

```{r btw-package, eval=FALSE}
# Overview of recipes package
btw(btw_tool_docs_package_help_topics("recipes"))

# Overview of parsnip
btw(btw_tool_docs_package_help_topics("parsnip"))

# Get vignettes
btw(vignette("workflows"))
btw(vignette("tuning", "tune"))
```

## Interactive Learning with AI

```{r btw-ai-chat, eval=FALSE}
library(ellmer)

# Create chat session
chat <- chat_anthropic() # or chat_ollama()

# Ask questions with context
chat$chat(
  btw(
    tune_grid,
    ames_folds,
    "I'm getting an error when tuning. How do I fix it?"
  )
)

# Get examples
chat$chat(
  btw(
    recipe,
    ames_train,
    "Show me 3 different ways to handle missing data in recipes"
  )
)
```

---

# 10. Best Practices

## Data Leakage Prevention

```{r data-leakage}
# WRONG: Computing statistics on full data
# mean_price <- mean(ames$Sale_Price)
# ames_normalized <- ames %>% mutate(Sale_Price = Sale_Price / mean_price)

# RIGHT: Let recipe handle preprocessing
# Statistics are computed only on training data
# and applied to test data
correct_recipe <- recipe(Sale_Price ~ ., data = ames_train) %>%
  step_normalize(all_numeric_predictors()) # Uses training means/SDs only
```

## Workflow Organization

```{r workflow-org, eval=FALSE}
# Good practice: Organize your modeling project

# 1. Data preparation script
# data_prep.R

# 2. Feature engineering and recipe
# recipe_definitions.R

# 3. Model specifications
# model_specs.R

# 4. Training script
# train_models.R

# 5. Evaluation script
# evaluate_models.R

# 6. Deployment script
# deploy_model.R
```

## Model Comparison

```{r model-comparison, eval=FALSE}
# Compare multiple models systematically
library(workflowsets)

# Define recipes
recipe_basic <- recipe(Sale_Price ~ ., data = ames_train) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())

recipe_engineered <- recipe(Sale_Price ~ ., data = ames_train) %>%
  step_log(Sale_Price) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_interact(~ Gr_Liv_Area:Year_Built)

# Define models
models <- list(
  lm = linear_reg() %>% set_engine("lm"),
  rf = rand_forest() %>% set_engine("ranger") %>% set_mode("regression"),
  xgb = boost_tree() %>% set_engine("xgboost") %>% set_mode("regression")
)

# Create all combinations
all_workflows <- workflow_set(
  preproc = list(basic = recipe_basic, engineered = recipe_engineered),
  models = models,
  cross = TRUE
)

# Fit all workflows
all_results <- all_workflows %>%
  workflow_map(
    fn = "fit_resamples",
    resamples = ames_folds,
    metrics = metric_set(rmse, rsq)
  )

# Compare results
all_results %>%
  rank_results(rank_metric = "rmse") %>%
  filter(.metric == "rmse")

# Visualize
all_results %>%
  autoplot()
```

---

# Resources

## Documentation
- [tidymodels.org](https://www.tidymodels.org/) - Official documentation
- [Tidy Modeling with R](https://www.tmwr.org/) - Free online book
- [recipes reference](https://recipes.tidymodels.org/reference/index.html)
- [parsnip models](https://www.tidymodels.org/find/parsnip/)

## Learning with btw
```{r btw-resources, eval=FALSE}
# Get started with btw
btw::btw() # Describe current workspace

# Learn any tidymodels function
btw::btw(your_function_here, your_data, "your question")

# Use with AI chat
library(ellmer)
chat <- chat_anthropic()
chat$chat(btw(function_name, data, "question"))
```

## Community
- [RStudio Community](https://community.rstudio.com/) - Ask questions
- [tidymodels GitHub](https://github.com/tidymodels) - Source code and issues
- [#tidymodels on Twitter](https://twitter.com/hashtag/tidymodels)

## Comparisons and Benchmarks
- [Caret vs. tidymodels](https://medium.com/data-science/caret-vs-tidymodels-create-complete-reusable-machine-learning-workflows-5c50a7befd2d) - Real-world comparison by Hannah Roos
- [From caret to tidymodels](https://www.tidyverse.org/blog/2020/04/tidymodels-org/) - Official migration guide

---

# Quick Reference

## Common Workflow Pattern

```{r quick-reference, eval=FALSE}
# 1. Split
split <- initial_split(data, prop = 0.75)
train <- training(split)
test <- testing(split)
folds <- vfold_cv(train, v = 10)

# 2. Recipe
rec <- recipe(outcome ~ ., data = train) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())

# 3. Model
model <- rand_forest(mtry = tune(), min_n = tune()) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# 4. Workflow
wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(model)

# 5. Tune
results <- wf %>%
  tune_grid(resamples = folds, grid = 10)

# 6. Finalize
best <- select_best(results, "rmse")
final_wf <- finalize_workflow(wf, best)
final_fit <- last_fit(final_wf, split)

# 7. Evaluate
collect_metrics(final_fit)
```

---

**Happy modeling with tidymodels!**
